---
title: "Package Test Script"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: html_document
runtime: shiny
---

<hr style="border-color: black; border-width: 2px;">
<br>
<h4> Name: `r your_name` </h4>
<h4> Email Address: `r your_email_address` </h4><br>

<hr style="border-color: black; border-width: 2px;">

<h3>Check packages are loaded:</h3>

```{r, echo=FALSE, results='asis'}
packages <- c("htmltools", "shiny", "shinyWidgets", "shinydashboard", "leaflet", "RColorBrewer",
              "lubridate", "dplyr", "rgdal", "raster", "rmarkdown", "png", "kableExtra",
              "rgeos", "magrittr", "mapview", "ggplot2")
for(i in packages){
  lapply(i, require, character.only = TRUE)
  pr = Xmisc::is.package.loaded(i, character.only = TRUE)
  if(pr==FALSE){
    pr = htmltools::HTML(paste0("<span style='color:red;'>", pr, "</span><br>"))
  } else {
    pr = htmltools::HTML(paste0("<span style='color:black;'>", pr, "</span><br>"))
  }
  cat(paste0(i, ": ", pr))
}
```

<hr style="border-color: black; border-width: 2px;">

<h3> Additional checks: </h3>

<li><b>ggplot2</b> barplot subsetted with <b>dplyr</b> from 2000 onwards (normally 1975), coloured red, blue & green by storm type with <b>RColorBrewer</b>.</li><br>
```{r echo=FALSE, fig.width=12, fig.height=5}
data("storms")

# Get colour palette
pal <- brewer.pal(name="Set1", n=3)

storms_sub <- storms %>% filter(year>=2000)
ggplot() + 
  geom_bar(data=storms_sub, aes(x=year, fill=status)) + 
  scale_fill_manual(values=pal) + 
  theme_classic()
```

<br>
<li><b>leaflet</b> map with points coloured white > red with <b>RColorBrewer</b> and labelled with <b>htmltools</b>, with Scotland polygon coloured orange, loaded and transformed using <b>rgdal</b>.</li><br>
```{r echo=FALSE, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
# Download cities shapefile
dsn <- system.file("vectors", package = "rgdal")[1]
scot <- readOGR(dsn=dsn, layer="scot_BNG", verbose=FALSE)
scot <- spTransform(scot, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))

# Set category as factor
storms$category <- factor(storms$category)

# etup colour palette
disc_pal <- brewer.pal(name="Reds", n=7)

# Create labels
map_labels <- lapply(seq(nrow(storms)), function(i) {
    paste0( '<p>', "Sent in by: ", '<b>', storms[i, "name"], '</b>')})

# Create plot
leaflet() %>% 
  addProviderTiles("Stamen.Terrain") %>%
  addPolygons(data=scot, fill="yellow", color="orange", weight=1) %>%
  addCircleMarkers(data=storms, radius=2, color = ~disc_pal, fillOpacity=1, 
                   label=lapply(map_labels, HTML), stroke=FALSE)
```

<br>
<li><b>kableExtra</b> formatted table ordered by date column, created using <b>lubridate</b>.</li><br>
```{r echo=FALSE, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
storms$date <- paste0(storms$day,"/", storms$month, "/", storms$year)
storms$date <- dmy(storms$date)

storms_summary <- storms %>%
  group_by(date, status, category) %>%
  summarise(n=length(date))
kable(storms_summary) %>%
  kable_styling(bootstrap_options = "condensed") %>%
  scroll_box(width = "100%", height = "400px")
```

<br><br><br>

