---
title: "Package Test Script"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: html_document
---

<hr style="border-color: black; border-width: 2px;">
<br>
<h4> Name: `r your_name` </h4>
<h4> Email Address: `r your_email_address` </h4><br>

<hr style="border-color: black; border-width: 2px;">

<h3>Check packages are loaded:</h3>

```{r, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
# Define a function to test if a padckage is installed
is.installed <- function(mypkg){
  is.element(mypkg, installed.packages()[,1])
}

# Set vector of package names
list.of.packages <- c("shiny", "shinyWidgets", "shinydashboard", "leaflet", "dplyr", 
                      "ggplot2", "RColorBrewer", "kableExtra",  "rgdal", "raster", "htmltools",
                      "rmarkdown", "png",  "rgeos", "magrittr", "mapview", "knitr",
                      "lubridate", "Xmisc", "pillar", "vctrs", "sf") 

# Print if package will load in correctly
for(i in list.of.packages){
  lapply(i, require, character.only = TRUE)
  pr = Xmisc::is.package.loaded(i, character.only = TRUE)
  if(pr==FALSE){
    pr = htmltools::HTML(paste0("<span style='color:red;'>", pr, "</span><br>"))
  } else {
    pr = htmltools::HTML(paste0("<span style='color:black;'>", pr, "</span><br>"))
  }
  cat(paste0(i, ": ", pr))
}

# Download dataset
data("iris")

# Add extra columns for date setup and long/lat
iris$year <- as.numeric(sample(x=1990:2015, size=nrow(iris), replace=TRUE))
iris$month <- as.numeric(sample(x=1:12, size=nrow(iris), replace=TRUE))
iris$day <- as.numeric(sample(x=1:28, size=nrow(iris), replace=TRUE))
iris$long <- sample(x=50:100, size=nrow(iris), replace=TRUE)
iris$lat <- sample(x=25:50, size=nrow(iris), replace=TRUE)
```

<hr style="border-color: black; border-width: 2px;">

<h3> Additional checks: </h3>

<b>ggplot2</b> barplot subsetted with <b>dplyr</b> from 2000 onwards, coloured red, blue & green by species with <b>RColorBrewer</b>.
<li>If If data is shown from 1990 to 2015, <b>dplyr</b> is missing.</li>
<li>If bars are plotted in grey, <b>rColorBrewer</b> is missing.</li>
<li>Check plot title for ggplot2/base R</li><br>
```{r echo=FALSE, fig.width=12, fig.height=5}
# Get colour palette from RColorBrewer, else set greyscale palette
if(Xmisc::is.package.loaded("RColorBrewer")){
  pal <- brewer.pal(name="Set1", n=3)
} else {
  pal <- c("darkgrey", "dimgrey", "grey")
}

# Subset data with dplyr, else use whole dataset
if(Xmisc::is.package.loaded("dplyr")){
  iris_sub <- iris %>% filter(year>=2000)
} else {
  iris_sub <- iris 
}

# Create plot with ggplot, else us base R
if(Xmisc::is.package.loaded("ggplot2")){
  ggplot() +
    geom_bar(data=iris_sub, aes(x=year, fill=Species)) +
    scale_fill_manual(values=pal) +
    theme_classic() + 
    ggtitle("Created with ggplot2")
} else {
  brks = seq(min(iris_sub$year), max(iris_sub$year), by=1)
  hist(iris_sub$year, main="Created with base R", breaks=brks, xlab="year", ylab="count", col=pal[1])
  hist(iris_sub$year[which(iris_sub$Species!="setosa")], breaks=brks, col=pal[2], add=TRUE)
  hist(iris_sub$year[which(iris_sub$Species=="virginica")], breaks=brks, col=pal[3], add=TRUE)
}

```

<br>
<b>leaflet</b> map with points coloured red and labelled with <b>htmltools</b>, with Scotland polygon coloured orange, loaded and transformed using <b>rgdal</b>.
<li>If data is plotted as a scatter plot, <b>leaflet</b> is missing. </li>
<li>If no polygons appear over Scotland, <b>rgdal</b> is missing. </li>
<li>If no labels appear, <htmltools</li> is missing. </li>

```{r echo=FALSE, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
# Download cities shapefile
if(Xmisc::is.package.loaded("rgdal")){
  dsn <- system.file("vectors", package = "rgdal")[1]
  scot <- readOGR(dsn=dsn, layer="scot_BNG", verbose=FALSE)
  scot <- spTransform(scot, CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
}

if(Xmisc::is.package.loaded("htmltools")){
map_labels <- lapply(seq(nrow(iris)), function(i) {
    paste0( '<p>', "Sepal Length (mm): ", '<b>', iris[i, "Sepal.Length"], '</b>')})
}

# Set category as factor
iris$Species <- factor(iris$Species)

# Create plot
if(Xmisc::is.package.loaded("leaflet")){
  map <- leaflet() %>%
  addProviderTiles("Stamen.Terrain") 
  
  if(Xmisc::is.package.loaded("htmltools")){
  map <- map %>%
    addCircleMarkers(data=iris, radius=4, color = "red", fillOpacity=1, stroke=FALSE,
                   label=lapply(map_labels, HTML))
  } else {
    map <- map %>%
    addCircleMarkers(data=iris, radius=4, color = "red", fillOpacity=1, stroke=FALSE)
  }
  
    if(Xmisc::is.package.loaded("rgdal")){
    map <- map %>%
      addPolygons(data=scot, fill="yellow", color="orange", weight=1, fillOpacity=1)
    }
  map
} else {
  plot(iris$long, iris$lat, col="red", pch=19)
}

```

<br>
<b>kableExtra</b> formatted table ordered by date column, created using <b>lubridate</b>.
<li>If table is not contained in a scroll box, <b>kableExtra</b> is missing. </li>
<li>If table is not ordered by date is in the format yyyy-mm-dd, <b>lubridate</b> is missing. </li><br>
```{r echo=FALSE, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
# Create date column, formatted as character
iris$date <- paste0(iris$day,"/", iris$month, "/", iris$year)

# Transform to date object
if(Xmisc::is.package.loaded("lubridate")){
  iris$date <- dmy(iris$date)
}

# Subset data
iris_subset <- iris[c("date", "Species", "Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")]

# Order by date
iris_subset <- iris_subset[order(iris_subset$date),]

# Create table
tbl <- knitr::kable(iris_subset, row.names=FALSE) 
if(Xmisc::is.package.loaded("kableExtra")){
  tbl <- kable_styling(kable_input=tbl, bootstrap_options = "condensed") %>%
  scroll_box(width = "100%", height = "400px")
}
tbl
```

<br><br><br>

