---
title: "Data Visualisation using RShiny: Day 1"
author: "Getting to know your data"
date: "1.1 Guide"
output: 
  html_document:
    theme: yeti
    highlight: tango
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(rgdal)
library(RColorBrewer)
library(kableExtra)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- readOGR("data/TZ_Region_2012", "TZ_Region_2012")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D")
```

<style>
h1, h2, h3, h4 {font-weight: bold;}
.question {font-weight: bold; color: #4E23B8;}
hr {border: 2px solid black;}
a:link {color: white; text-decoration: none;}
a:visited {color: white; text-decoration: none;}
a:hover {color: #c9bde9; text-decoration: none;}
a:active {color: white; text-decoration: none;}
.panel-aims {color: white; background-color: #4E23B8; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
.panel-section {color: black; background-color: #c9bde9; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
</style>

<!--- ---------------------------------------------------------------------- -->
<!--- Aims section -->

<br>
<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-aims"> <!--- Begin panel -->
<h3>Contents</h3>
Click to view a specific section of the document.  
<br>
<ul>
<li><a href="#1">Loading in packages</a></li>
<li><a href="#2">Loading in data</a></li>
<li><a href="#3">Exploring data</a></li>
</ul>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<!--- ---------------------------------------------------------------------- -->
<!--- 1. Loading in packages -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1">Loading in packages</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>At the very beginning of any R script, you should load in any packages you plan on using.<.li>
<li>Packages contain functions to complete a <em>very</em> broad range of processes. </li>
<li>To load in your packages, we can use the ``library()`` function:</li>

``` {r, eval=FALSE}
library(dplyr)
```
</ul>
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2. Loading in data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2">Loading in data</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>As with packages, you should load in any data you want to use at the beginning on your R script.</li>
<li>This allows you to easily see what data is being loaded, rather than having to search through your code.</li>
<li>There are multiple functions to load in your data - we will only be using a <em>comma-seperated values</em> spreadsheet, so we need the ``read.csv()`` function:

``` {r, eval=FALSE}
raw_data <- read.csv("data/raw_data.csv") 
```
</ul>
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 3. Exploring data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="3">Exploring data</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>We have several functions to get to know the data: 
<br>

``` {r, echo=FALSE}
explore <- data.frame("function name"=c("``View()``", "``nrow()``", "``names()``", "``dim()``", 
                                   "``summary()``", "``head()``", "``tail()``", "``str()``",
                                   "``unique()``", "``table()``", "``?``", "``??``"),
                      "description"=c("view the dataframe as a spreadsheet",
                                      "report the number of rows in the data frame",
                                      "report the column name in the data frame",
                                      "report the dimensions of the data frame",
                                      "report the summary statistics (counts, min, median, mean, max)",
                                      "print the first 6 rows of the data",
                                      "print the last 6 rows of the data",
                                      "report the variable type in each column (e.g. Factor, num (number), int (integer))",
                                      "report the unique elements of a variable",
                                      "calculate the count number for variables in the column(s) specified",
                                      "open the helpfile for a function contained in a loaded package",
                                      "open the helpfile for a function contained in an package that isn't loaded"))
kable_styling(kable(explore, col.names=c("function name", "description")), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```

</li>
</ul>
<br><br>