---
title: "Data Visualisation using RShiny: Day 1"
author: "Data subsetting and summarising"
date: "1.2 Guide"
output: 
  html_document:
    theme: yeti
    highlight: tango
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(rgdal)
library(kableExtra)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- readOGR("data/TZ_Region_2012", "TZ_Region_2012")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D")
```

<style>
h1, h2, h3, h4 {font-weight: bold;}
.question {font-weight: bold; color: #4E23B8;}
hr {border: 2px solid black;}
a:link {color: white; text-decoration: none;}
a:visited {color: white; text-decoration: none;}
a:hover {color: #c9bde9; text-decoration: none;}
a:active {color: white; text-decoration: none;}
.panel-aims {color: white; background-color: #4E23B8; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
.panel-section {color: black; background-color: #c9bde9; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
</style>

<!--- ---------------------------------------------------------------------- -->
<!--- Aims section -->

<br>
<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-aims"> <!--- Begin panel -->
<h3>Contents</h3>
Click to view a specific section of the document.  
<br>
<ul>
<li><a href="#1">The dplyr package</a></li>
<li><a href="#2">Subsetting</a></li>
<li><a href="#3">Summarising</a></li>
<li><a href="#4">Adding new variables</a></li>
<li><a href="#5">The ``%`` (pipe) function</a></li>
</ul>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<!--- ---------------------------------------------------------------------- -->
<!--- 1. The dplyr package -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1">The dplyr package</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

The dplyr package contains some very helpful functions to manipulate data, including subsetting and summarising:  
<br>
``` {r, echo=FALSE}
explore <- data.frame("function name"=c("``select()``", "``filter()``", "``summarise()``", "``group_by()``", "``mutate()``"),
                      "description"=c("subsets columns based on their names",
                                      "subsets rows based on their values",
                                      "calculates summary statistics",
                                      "groups variables for summarising",
                                      "adds new columns that are functions of existing variables"))
kable_styling(kable(explore, col.names=c("function name", "description")), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2. Subsetting -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2">Subsetting</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4>``select()``</h4>
<ul>
<li>The `` select()`` function is used to select columns of data by their column name.</li>  
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
select(raw_data, date, region, x, y, species, density)
```

``` {r, echo=FALSE}
select_1 <- data.frame(select(raw_data, date, region, x, y, species, density))
kable_styling(kable(select_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed")) %>%
  scroll_box(height="200px")
```

<br><br>
<h4>``filter()``</h4>
<ul>
<li>The `` filter()`` function is used to select rows based on their values.</li>
<li>To select the observations you want, it is useful to know some comparison operators:</li>
</ul>
``` {r, echo=FALSE}
explore <- data.frame("operator"=c(">", ">=", "!=", "==", "&", 
                                   "|", "!", "``%in% c(...)``"),
                      "description"=c("greater than", "greater than or equal to", "not equal",
                                      "equal", "and", "or", "not", "one in a list of elements"))
kable_styling(kable(explore, escape=FALSE), 
              font_size=14, full_width=FALSE, position="left", 
              bootstrap_options=c("bordered", "condensed"), protect_latex = TRUE)
```

<em>E.g.</em>
```{r, eval=FALSE}
filter(raw_data, region == "Mara" & age <= 5)
```

``` {r, echo=FALSE}
filter_1 <- data.frame(filter(raw_data, region == "Mara" & age <= 5))
kable_styling(kable(filter_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed")) %>%
  scroll_box(height="200px")
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 3. Summarising -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="3">Summarising</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4> ``summarise()``</h4>
<ul>
<li>The `` summarise()`` function is used to calculate summaries of variables in your data.</li>
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
summarise(raw_data, total_cases = n(), mean_age = mean(age))
```

``` {r, echo=FALSE}
summarise_1 <- data.frame(summarise(raw_data, total_cases = n(), mean_age = mean(age)))
kable_styling(kable(summarise_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```

<br>
<h4> ``group_by()``</h4>
<ul>
<li>The `` group_by `` function is used to group variables before summarising.</li>
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
grouped_data <- group_by(raw_data, species)
summarise(grouped_data, total_cases = n(), mean_age = mean(age))
```

``` {r, echo=FALSE}
grouped_1 <- group_by(raw_data, species)
summarise_2 <- data.frame(summarise(grouped_1, total_cases = n(), mean_age = mean(age)))
kable_styling(kable(summarise_2), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 4. Adding new variables -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="4">Adding new variables</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4>``mutate()``</h4>
<ul>
<li>The ``mutate()`` function adds new columns that are functions of existing variables.</li>
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
raw_data <- mutate(raw_data, animal_type = ifelse(species == "human", "human", "animal"))
```

``` {r, echo=FALSE}
mutate_1 <- mutate(raw_data, animal_type = ifelse(species == "human", "human", "animal"))
kable_styling(kable(mutate_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed")) %>%
  scroll_box(height="200px")
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 5. The ``%`` (pipe) function -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="5">The ``%`` (pipe) function</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>dplyr functions can be used individually, or can be 'stacked' using the ``%`` (pipe) function:
<br>

```{r, eval=FALSE}
df_filtered <- filter(raw_data, species != "human") 
df_grouped <- group_by(df_filtered, animal_type)
df_summarised <- summarise(df_grouped, n())

df_summarised <-  raw_data %>% 
  filter(species != "human") %>% 
  group_by(animal_type) %>% 
  summarise(n())
```
</li>
<li>Using the pipe function makes the computing more efficient as it passes the output of one function directly into the next.</li>
<li>This reduces the number of objects you save to the global environment.</li>
</ul>
<br>
