---
title: "Data Visualisation using RShiny: Day 1"
author: "Building exploratory plots"
date: "1.3 Guide"
output: 
  html_document:
    theme: yeti
    highlight: tango
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(rgdal)
library(kableExtra)
library(stats)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- readOGR("data/TZ_Region_2012", "TZ_Region_2012")
mara_shp <- region_shp[which(region_shp$Region_Nam=="Mara"),]

region_shp$id <- rownames(region_shp@data)
region_shp_df <- fortify(region_shp, region = "id")
region_shp_df <- left_join(region_shp_df, region_shp@data, by = "id")

mara_shp$id <- rownames(mara_shp@data)
mara_shp_df <- fortify(mara_shp, region = "id")
mara_shp_df <- left_join(mara_shp_df, mara_shp@data, by = "id")

summary_data <- raw_data %>%
  filter(species=="human") %>%
  group_by(month, date, sex) %>%
  summarise(mean_age = mean(age),
            mean_density = mean(density)) %>%
  arrange(-mean_age)

summary_data_means <- raw_data %>%
  group_by(species) %>%
  summarise(mean_age = round(mean(age), digits=1),
            sd_age = round(sd(age), digits=1))

movement <- raw_data %>%
  group_by(month, species, region) %>%
  select(region, month, x,y) %>%
  filter(species == "dog" & region=="Mara")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D")
```

<style>
h1, h2, h3, h4 {font-weight: bold;}
.question {font-weight: bold; color: #4E23B8;}
.blue {color: #4E23B8;}
hr {border: 2px solid black;}
a.title:link {color: white; text-decoration: none;}
a.title:visited {color: white; text-decoration: none;}
a.title:hover {color: #c9bde9; text-decoration: none;}
a.title:active {color: white; text-decoration: none;}
a.main:link {color: #4E23B8;}
a.main:visited {color: #4E23B8;}
a.main:hover {color: #4E23B8; font-weight: bold;}
a.main:active {color: #4E23B8;}
.panel-aims {color: white; background-color: #4E23B8; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
.panel-section {color: black; background-color: #c9bde9; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
</style>

<!--- ---------------------------------------------------------------------- -->
<!--- Aims section -->

<br>
<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-aims"> <!--- Begin panel -->
<h3>Contents</h3>
Click to view a specific section of the document.  
<br>
<ul>
<li><a class="title" href="#1">The ggplot2 package</a></li>
<li><a class="title" href="#2">Setting up a basic plot</a></li>
<li><a class="title" href="#3">Customising plot appearence</a></li>
<li><a class="title" href="#4">Plot type examples</a></li>
<li><a class="title" href="#5">More Information</a></li>
</ul>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<!--- ---------------------------------------------------------------------- -->
<!--- 1. The ggplot2 package -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1">The ggplot2 package</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

The ggplot2 package has a range of functions to allow easy and buildable plotting. It is designed so that all plotting functions take similar arguments, allowing users to quickly and easily build quite complicated plots.   
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2. Setting up a basic plot -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2">Setting up a basic plot</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

Plots in ggplot2 follow a simple basic structure. Here is an example using a barplot:  
<br>
<h4>Step 1: Begin the plot</h4>
<b class="blue">ggplot() + </b>  
<br>

<h4>Step 2: add a plot type</h4>
ggplot() + <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; <b class="blue">geom_bar</b>()  
<br>

<h4>Step 3: add the data</h4> 
ggplot() + <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; geom_bar(<b class="blue">data = raw_data</b>)  
<b>or</b>
<br>
ggplot(<b class="blue">data = raw_data</b>) + <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; geom_bar()  
<br>

<h4>Step 4: specify the variables you want mapped</h4>
Here, we enclose mapping variables within the ``aes()`` function. This tells the plot what to show on the x-axis, y-axis etc.
ggplot() + <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; geom_bar(data = raw_data, <b class="blue">aes(x = species)</b>)  
<br>

<h4>Complete code:</h4> 

```{r, echo = T, fig.height = 3}
ggplot() + 
  geom_bar(data = raw_data, aes(x = species))
```
<br><br>

<!--- ---------------------------------------------------------------------- -->
<!--- 3. Customising plot appearence -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="3">Customising plot appearence</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4> Changing plot colour </h4>
<ul>
<li>You can change the overall colour with a named colour or a hex colour (<em>e.g. "#4E23B8"</em>)</li>
<li>This can be ``fill`` or ``color`` depending on the plot type:
<br>
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species), fill = "blue")  
```
</li>
<br>
<li>Or you can set the fill colour to change with a variable:
<br>
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species, fill = species))  
```
</li>
<br>
<li>The variable colour can be different to what is plotted on the x-axis:
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species, fill = sex))  
```
</li>
</ul>
<br>

<h4> Setting the plot colours that are mapped to variables </h4>
<ul>
<li>To set the plot colour, you have to use either ``scale_fill_...`` or ``scale_color_...`` depending on the plot type:
<li>The RColorBrewer package contains a <a class="main" href="https://www.r-graph-gallery.com/38-rcolorbrewers-palettes/">list</a> of freely available palettes you can use:
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species, fill = sex)) + 
  scale_fill_brewer(palette = "Set1")
```
</li>
<br>
<li>Alternatively, you can provide a vector of your own colour choices:
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species, fill = sex)) + 
  scale_fill_manual(values = c("blue", "red"))
```
</li>
</ul>
<br><br>

<h4> Adding axis labels </h4>
<ul>
<li>You can add axis labels easily with the ``xlab()``, ``ylab()`` or ``labs()`` functions: 

```{r, echo = T, eval = F}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species), fill = "blue") + 
  xlab("Species") + 
  ylab("Cases")
```
<b>or</b>
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species), fill = "blue") + 
  labs(x = "Species", y = "Cases")
```
</li>
</ul>
<br>
<h4> Adding a plot title </h4>
<ul>
<li>You can add a plot title and/or subtitle as part of the ``labs()`` function:
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species), fill = "blue") + 
  labs(x = "Species", y = "Cases",
       title = "Number of cases observed by species", subtitle = "Cases observed in Tanzania")
```
</li>
</ul>
<br>
<h4> Adding a theme </h4>
<ul>
<li>Themes can be a quick way to alter the appearence of your plot - our favourite is ``theme_classic()``:
```{r, echo = T, fig.height = 3}
ggplot() +  
  geom_bar(data = raw_data, aes(x = species), fill = "blue") + 
  labs(x = "Species", y = "Cases",
       title = "Number of cases observed by species", subtitle = "Cases observed in Tanzania") + 
  theme_classic()
```
</li>
</ul>
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 4. Plot type examples -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="4">Plot type examples</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>ggplot2 has individual functions for each type of plot</li>
<li>There are also functions for additional plotting features such as lines, error bars and text.</li>
<li>This allows the user to layer multiple plots easily e.g. bars and lines.</li>
<li>We have provided some examples of different plots below:</li>
</ul>
<br>

<h3 class = "blue">Common plots</h3>
<div class = "row">
<div class = "col-md-6">
<h4>geom_bar</h4>
Barplot, ggplot2 automatically counts totals provides by x<br>
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_bar(data=raw_data, 
           aes(x=species)) + 
  theme_classic()
```
</div>
<div class = "row">
<div class = "col-md-6">
<h4>geom_col</h4>
Barplot, the user provides summary data for x and y <br>
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_col(data=summary_data_means, 
           aes(x=species, y=mean_age)) + 
  theme_classic()
```
</div>
</div>

<div class = "row">
<div class = "col-md-6">
<h4>geom_histogram</h4>
Histogram, ggplot2 automatically counts totals along a continuous gradient <br>
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_histogram(data=raw_data, 
                 aes(x=month)) + 
  theme_classic()
```
</div>
<div class = "row">
<div class = "col-md-6">
<h4>geom_point</h4>
Point/scatter, user provides x and y variables <br>
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_point(data=raw_data, aes(x=x, y=density)) +
  theme_classic()
```
</div>
</div>

<div class = "row">
<div class = "col-md-6">
<h4>geom_line</h4>
Lineplot, connects observations in the order they appear on the x-axis (e.g. over time)
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_line(data=summary_data, 
            aes(x=month, y=mean_age, color=sex)) + 
  theme_classic()
```
</div>
<div class = "row">
<div class = "col-md-6">
<h4>geom_path</h4>
Lineplot, connects observation in the order they appear in the data (e.g. observation number)
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_path(data=movement, 
            aes(x=x, y=y), color="red") + 
  labs(x="Longitude", y="Latitude") + 
  theme_classic() 
```
</div>
</div>

<br>
<h3 class = "blue">Additional plots</h3>
<div class = "row">
<div class = "col-md-6">
<h4>geom_boxplot</h4>
Boxplot, draws boxes and outliers, user sets x and y
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_boxplot(data=raw_data, 
             aes(x=species, y=age)) + 
  theme_classic()
```
</div>
<div class = "row">
<div class = "col-md-6">
<h4>geom_polygon</h4>
Spatial polygons, data must be formatted as a data frame
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_polygon(data=region_shp_df, 
               aes(x=long, y=lat, group=group), 
               fill="white", color="black") + 
  theme_classic() + coord_equal()
```
</div>
</div>

<br>
<h3 class = "blue">Layering plots</h3>
<div class = "row">
<div class = "col-md-6">
<h4>geom_polygon + geom_path</h4>
To layer multiple plots to one output, simply join with a ``+`` symbol
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot() + 
  geom_polygon(data=mara_shp_df, 
               aes(x=long, y=lat, group=group), 
               fill="white", color="black") + 
  geom_path(data=movement, 
            aes(x=x, y=y), color="red") + 
  theme_classic() + coord_equal()
```
</div>
</div>

<br>
<h3 class = "blue">Additional features</h3>
<div class = "row">
<div class = "col-md-6">
<h4>geom_abline, geom_hline and geom_vline</h4>
Diagonal, horizontal and vertical line plotting
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
coef(lm(density~x, data=raw_data))
ggplot() + 
  geom_point(data=raw_data, aes(x=x, y=density)) + 
  geom_abline(intercept=-5800, slope=190, 
              color="red", size=1) + 
  geom_hline(yintercept=9000, linetype="dashed") +
  geom_vline(xintercept=35, color="blue", size=1.5) +
  theme_classic()
```
</div>
<div class = "row">
<div class = "col-md-6">
<h4>geom_errorbar</h4>
Error bars matched to plot variables
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot(data=summary_data_means) + 
  geom_col(aes(x=species, y=mean_age, fill=species)) + 
  geom_errorbar(aes(x=species, 
                    ymin=mean_age-sd_age,
                    ymax=mean_age+sd_age),
                width=0.2) + 
  theme_classic()
```
</div>
</div>

<div class = "row">
<div class = "col-md-6">
<h4>geom_text and geom_label</h4>
Plot text from data frame with (geom_label) or without (geom_text) white background
```{r echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3}
ggplot(data=summary_data_means) + 
  geom_col(aes(x=species, y=mean_age)) + 
  geom_text(aes(x=species, y=mean_age+5, 
                label=paste0("sd ", sd_age))) + 
  geom_label(aes(x=species, y=mean_age+10, 
                label=paste0("mean ", mean_age))) + 
  theme_classic() + 
  theme(panel.background = element_rect(fill="#fcf0cd"))
```
</div>
</div>

<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 5. More Information -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="5">More Information</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

There is <b>MUCH</b> more to ggplot2 than we have desccribed here!  
For more information on plot types and formatting plot appearence, see <a class="main"  href="https://ggplot2.tidyverse.org/reference/" target="_blank">here</a>.
<br><br>
