---
title: "Data Visualisation using RShiny: Day 2"
author: "Intro to leaflet"
date: "2.2 Guide"
output: 
  html_document:
    theme: yeti
    highlight: tango
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(leaflet)
library(rgdal)
# library(kableExtra)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- readOGR("data/TZ_Region_2012", "TZ_Region_2012")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D")
```

<style>
h1, h2, h3, h4 {font-weight: bold;}
.question {font-weight: bold; color: #4E23B8;}
hr {border: 2px solid black;}
a:link {color: white; text-decoration: none;}
a:visited {color: white; text-decoration: none;}
a:hover {color: #c9bde9; text-decoration: none;}
a:active {color: white; text-decoration: none;}
.panel-aims {color: white; background-color: #4E23B8; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
.panel-section {color: black; background-color: #c9bde9; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
</style>

<!--- ---------------------------------------------------------------------- -->
<!--- Contents section -->

<br>
<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-aims"> <!--- Begin panel -->
<h3>Contents</h3>
Click to view a specific section of the document.  
<br>
<ul>
<li><a href="#1">Loading and plotting the data</a></li>
<li><a href="#2">Options for tiles</a></li>
<li><a href="#3">Options for plotting point data</a></li>
<li><a href="#4">Colouring by discrete variables</a></li>
<li><a href="#5">Colouring by continuous variables</a></li>
</ul>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<!--- ---------------------------------------------------------------------- -->
<!--- 1. Loading and plotting the data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1">Loading and plotting the data</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

We'll start of plotting some point data associated with our hypothetical outbreak. Before visualising the data, we'll subset the data to select only observations from a single year and create a new variable that groups species into types.
<br>
``` {r, echo=T}

```
<br>
  
We use ``leaflet()`` to create a leaflet widget and can pipe this into a function such as ``addCircles()`` to get a basic plot of the data in space.
<br>
``` {r, eval = F, echo=T}
m <- leaflet() %>%
  addCircles(data = map_data, lng = ~x, lat = ~y)
```
<br>


<!--- ---------------------------------------------------------------------- -->
<!--- 2. Options for tiles -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2">Options for tiles</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

The default OpenStreetMap tiles are added to a leaflet widget using ``addTiles()`` while third party tiles can be added using ``addProviderTiles()``

``` {r, eval = F, echo=T}
# Initialise map centred around a location specified using latitude and longitude
m <- leaflet() %>%
  setView(lng = 38.9, lat = -6.4333, zoom = 12)

# addTiles() gives the default OpenStreetMap tiles
m %>% addTiles()

# Options for adding third party tiles "Stamen.Terrain"
m %>% addProviderTiles(provider = "Stamen.Terrain")
m %>% addProviderTiles(provider = providers$Stamen.Terrain)

# some other examples
m %>% addProviderTiles(provider = providers$Wikimedia)
m %>% addProviderTiles(provider = providers$Esri.NatGeoWorldMap)
m %>% addProviderTiles(provider = providers$Esri.WorldImagery)
m %>% addProviderTiles(provider = providers$OpenTopoMap)
```

<!--- ---------------------------------------------------------------------- -->
<!--- 3. Options for plotting point data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="3">Options for plotting point data</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

. Values for opacity below 1 introduce transparency to markers/lines etc. Popup info to display when a marker is clicked can be added directly but often is often neater to store this in advance. In popup info ``"<br>"`` produces a line break.
```{r, eval = F, echo = T}
popupInfo <- paste("Date: ", map_data$date, "<br>",
                   "Species: ", map_data$species, "<br>",
                   sep = " ")
m %>% addCircleMarkers(lng=~x, lat=~y, popup = popupInfo,
                       radius = 10,
                       stroke = T, # plot a 
                       fill.colour = "red" # central colour
                       fill.opacity = 0.6,
                       colour = "grey", # stroke colour
                       opacity = 0.9) # stroke opacity
```


<!--- ---------------------------------------------------------------------- -->
<!--- 4. Colouring by discrete variables -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="4">Colouring by discrete variables</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

Choose colours to form a colour pallete then use ``colorFactor()`` creates a function that can be used to colour layers such as circles:

```{r, eval = F, echo = T}
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(colour_pal, domain = c("Human", "Domestic", "Wildlife"))
addCircles(data = map_data, lng = ~x, lat = ~y,
           color = myPal(map_data$species_type), opacity = 0.9)
```

<!--- ---------------------------------------------------------------------- -->
<!--- 5. Colouring by continuous variables -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="5">Colouring by continuous variables</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

