---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "**Tanzania**, August, 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggmap)
library(leaflet)
library(rgdal)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- rgdal::readOGR("data/TZ_Region_2012_small/")

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3.5)
```

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
.blue {color: #4E23B8;}
.hl1-background {background-color: #fcf0cd;}
.hl2-background {background-color: #dfecfc;}

</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

  $('footnote').each(function(index) {
    var text  = $(this).html();
    var fnNum = (index+1).toString();
    $(this).html(fnNum.sup());

    var footnote   = fnNum + '. ' + text + '<br/>';
    var oldContent = $(this).parents('slide').children('div.footnotes').html();
    var newContent = oldContent + footnote;
    $(this).parents('slide').children('div.footnotes').html(newContent);
  });
});
</script>


<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Day 1 recap and intro to mapping in R </b></li>
> - <b class="blue"> Building a leaflet map in R </b></li>
> - Building an interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!

<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Day 1 recap and intro to mapping in R </b></li>
> - <b class="blue"> Building *an interactive* leaflet map in R </b></li>
> - Building an **even more** interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!


<!----------------------------------------------------------------------------->
# Recap day 1

<!----------------------------------------------------------------------------->
## Recap day 1

```{r, echo = T}
raw_data <- read.csv("data/raw_data.csv")

# str() gives the structure of the data
str(raw_data)
```

x = longitude and y = latitude in decimal degrees (not degrees, minutes and seconds)

<!----------------------------------------------------------------------------->
## The pipe (``%>%``) from the *magrittr* package 

What will the following code do?

```{r, eval = F}
raw_data %>%
  filter(species == "human") %>% 
  summarise(new_value = mean(age))
```

<!----------------------------------------------------------------------------->
## The pipe (``%>%``) from the *magrittr* package 

What will the following code do?

```{r, eval = T}
raw_data %>%
  filter(species == "human") %>% 
  summarise(new_value = mean(age))
```

<!----------------------------------------------------------------------------->
## The ``ifelse()`` function

What will the following line of code do?

```{r, eval = F}
ifelse("dog" %in% c("dog", "bird"), 2+2, "Hi")
```

<!----------------------------------------------------------------------------->
## The ``ifelse()`` function

What will the following line of code do?

```{r, eval = T}
ifelse("dog" %in% c("dog", "bird"), 2+2, "Hi")
```

<!----------------------------------------------------------------------------->
## ``ifelse()``

What will the following line of code do?

```{r, eval = T}
ifelse("dog" %in% c("dog", "bird"), 2+2, "Hi")
```

How about this one?

```{r, eval = F}
ifelse(3 > 2 & 7 < 5, "Hi", mean(raw_data$density))
```
<!----------------------------------------------------------------------------->
## ``ifelse()``

What will the following line of code do?

```{r, eval = T}
ifelse("dog" %in% c("dog", "bird"), 2+2, "Hi")
```

How about this one?

```{r, eval = T}
ifelse(3 > 2 & 7 < 5, "Hi", mean(raw_data$density))
```

<!----------------------------------------------------------------------------->
## Your Turn 2.1a

> - Navigate to Day 2/worksheets, and make a copy of the **R** script named <b>dat2_leaflet_intro.R</b> which you can work with.
<br><br><br>

Follow the code in this script to:

1) Load packages we'll need (``library()``) and the data set from yesterday
<br><br>
1) From "species", create a new column/variable 'species type' (Human, Domestic or Wildlife)
<br><br>
2) create ``map_data`` --- a dataframe consisting of 2014 data only

<!----------------------------------------------------------------------------->
## Data preparation

Creating a column/variable for 'species type'

```{r, echo = TRUE}
raw_data <- mutate(raw_data,
                   species_type = ifelse(species == "human", "Human",
                                         ifelse(species %in% c("cat", "dog"),
                                                "Domestic", "Wildlife")))


table(raw_data$species, raw_data$species_type)
```


<!----------------------------------------------------------------------------->
## Data preparation

Subsetting to only include 2014

```{r, echo = TRUE}
"2014-08-11" %>% substr(1, 4)

"Kilimanjaro" %>% substr(1, 4)

map_data <- raw_data %>% 
  mutate(year = substr(date, 1,4)) %>% 
  filter(year == 2014)
```

<!----------------------------------------------------------------------------->
# Simple static maps

<!----------------------------------------------------------------------------->
## Map Basics

The simplest map of this data using base ``plot()``

```{r, echo = TRUE, fig.height=4.4}
plot(map_data$y ~ map_data$x, asp = 1)
```

<!----------------------------------------------------------------------------->
## Shapefile data

A simple verion using ``ggplot()`` with points coloured by human population density

```{r, echo = TRUE, fig.height = 3.5}
ggplot(data = map_data, aes(x = x, y = y, col = density)) +
  geom_point(alpha = 0.4) + # makes points transparent
  coord_equal() # sets aspect ratio 1:1
```

<!----------------------------------------------------------------------------->
## **ggmap** and ``qmplot()`` {.smaller}

``qmplot()`` is the ggmap package's version of **ggplot2**'s ``ggplot()``

Using ``I()`` prevents the plotting of a legend beside the map

```{r, echo = T, fig.height = 3.5, message=FALSE}
qmplot(data = map_data, x = x, y = y, colour = I("red"),
       size = I(2), darken = .2, zoom = 4)
```
