---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "**Tanzania**, August, 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggmap)
library(leaflet)
library(rgdal)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- rgdal::readOGR("data/TZ_Region_2012_small/")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D") # 2:"#3E507F",
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
.blue {color: #4E23B8;}
.hl1-background {background-color: #fcf0cd;}
.hl2-background {background-color: #dfecfc;}

</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

  $('footnote').each(function(index) {
    var text  = $(this).html();
    var fnNum = (index+1).toString();
    $(this).html(fnNum.sup());

    var footnote   = fnNum + '. ' + text + '<br/>';
    var oldContent = $(this).parents('slide').children('div.footnotes').html();
    var newContent = oldContent + footnote;
    $(this).parents('slide').children('div.footnotes').html(newContent);
  });
});
</script>


<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Day 1 recap and intro to mapping in R </b></li>
> - <b class="blue"> Building a leaflet map in R </b></li>
> - Building an interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!

<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Day 1 recap and intro to mapping in R </b></li>
> - <b class="blue"> Building *an interactive* leaflet map in R </b></li>
> - Building an **even more** interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!


<!----------------------------------------------------------------------------->
## Recap day 1 {.smaller}

Let's remind ourselves what the data looked like

```{r, echo = T}
raw_data <- read.csv("data/raw_data.csv")

# str() gives the structure of the data
str(raw_data)
```

<!----------------------------------------------------------------------------->
## Recap day 1

Reading in data from shape file and converting to a dataframe

```{r, echo = T}
region_shp <- rgdal::readOGR("data/TZ_Region_2012", "TZ_Region_2012")
region_shp$id <- rownames(region_shp@data)
region_shp_df <- fortify(region_shp, region = "id")
region_shp_df <- left_join(region_shp_df, region_shp@data, by = "id")
```

<!----------------------------------------------------------------------------->
## Your Turn 2.1a

> - Navigate to Day 2/worksheets, and open <b>dat2_leaflet_intro.R</b>
<br><br>
> - Load libraries and datasets repeating code from previous slides
<br><br>
> - Follow code to create to:

1) create a new column/variable 'species type' (Human, Domestic or Wildlife)

2) create ``map_data`` --- a dataframe consisting of 2014 data only

<!----------------------------------------------------------------------------->
## Data preparation

Before we think about plotting, let's create a variable for 'species type'

```{r, echo = TRUE}
# create variable for type of species
raw_data$species_type <- ifelse(raw_data$species == "human", "Human", NA)
raw_data$species_type[which(raw_data$species %in% c("dog", "cat"))] <- "Domestic"
raw_data$species_type[which(raw_data$species %in% c("jackal", "lion"))] <- "Wildlife"

table(raw_data$species, raw_data$species_type)
```


<!----------------------------------------------------------------------------->
## Data preparation

Subsetting to only include 2014

```{r, echo = TRUE}
raw_data$date[1]
raw_data$date[1] %>% substr(1, 4)

map_data <- raw_data %>% 
  dplyr::mutate(year = substr(date, 1,4)) %>% 
  dplyr::filter(year == 2014)
```

<!----------------------------------------------------------------------------->
# Simple static maps

<!----------------------------------------------------------------------------->
## Map Basics

The simplest map of this data using base ``plot()``

```{r, echo = TRUE}
plot(map_data$y ~ map_data$x, asp = 1)
```

<!----------------------------------------------------------------------------->
## Map Basics

A simple verion using ``ggplot()`` with points coloured by human population density

```{r, echo = TRUE, fig.height = 4}
ggplot(data = map_data, aes(x = x, y = y, col = density)) +
  geom_point(alpha = 0.4) + # makes points transparent
  coord_equal() # sets aspect ratio 1:1
```

<!----------------------------------------------------------------------------->
## Map Basics {.smaller}

Combine previous plot with shape file data on regions

```{r, eval = T, echo = TRUE, fig.height = 3.5}
ggplot() +
  geom_polygon(data = region_shp_df, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = map_data, aes(x = x, y = y, col = density)) +
  coord_equal() # sets aspect ratio 1:1
```

<!----------------------------------------------------------------------------->
## **ggmap** and ``qmplot()`` {.smaller}

``qmplot()`` is the ggmap package's version of **ggplot2**'s ``ggplot()``

Using ``I()`` prevents the plotting of a legend beside the map

```{r, fig.height = 3.5, message=F}
qmplot(data = map_data, x = x, y = y, colour = I("red"),
       size = I(2), darken = .2, zoom = 5)
```


```{r, fig.height = 3.5, message=F}
qmplot(data = map_data, x = x, y = y, colour = I("red"),
       size = I(2), darken = .2, zoom = 5)
```
