---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "**Tanzania**, August, 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(leaflet)
library(rgdal)

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- rgdal::readOGR("data/TZ_Region_2012", "TZ_Region_2012")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D") # 2:"#3E507F",
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
.blue {color: #4E23B8;}
.hl1-background {background-color: #fcf0cd;}
.hl2-background {background-color: #dfecfc;}

</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

  $('footnote').each(function(index) {
    var text  = $(this).html();
    var fnNum = (index+1).toString();
    $(this).html(fnNum.sup());

    var footnote   = fnNum + '. ' + text + '<br/>';
    var oldContent = $(this).parents('slide').children('div.footnotes').html();
    var newContent = oldContent + footnote;
    $(this).parents('slide').children('div.footnotes').html(newContent);
  });
});
</script>


<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Day 1 recap and intro to mapping in R </b></li>
> - <b class="blue"> Building a leaflet map in R </b></li>
> - Building an interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!

<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Day 1 recap and intro to mapping in R </b></li>
> - <b class="blue"> Building *an interactive* leaflet map in R </b></li>
> - Building an **even more** interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!


<!----------------------------------------------------------------------------->
## Recap day 1 {.smaller}

Let's remind ourselves what the data looked like

```{r, echo = T}
raw_data <- read.csv("data/raw_data.csv")

# str() gives the structure of the data
str(raw_data)
```

<!----------------------------------------------------------------------------->
## Recap day 1

Reading in data from shape file and converting to a dataframe

```{r, echo = T}
region_shp <- rgdal::readOGR("data/TZ_Region_2012", "TZ_Region_2012")
region_shp$id <- rownames(region_shp@data)
region_shp_df <- fortify(region_shp, region = "id")
region_shp_df <- left_join(region_shp_df, region_shp@data, by = "id")
```

<!----------------------------------------------------------------------------->
## Your Turn

> - Navigate to Day 2/worksheets, and open <b>dat2_leaflet_intro.R</b>
<br><br>
> - Load libraries and datasets repeating code from previous slides
<br><br>
> - Follow code to create to:

1) create a new column/variable 'species type' (Human, Domestic or Wildlife)

2) create ``map_data`` --- a dataframe consisting of 2014 data only

<!----------------------------------------------------------------------------->
## Data preparation

Before we think about plotting, let's create a variable for 'species type'

```{r, echo = TRUE}
# create variable for type of species
raw_data$species_type <- ifelse(raw_data$species == "human", "Human", NA)
raw_data$species_type[which(raw_data$species %in% c("dog", "cat"))] <- "Domestic"
raw_data$species_type[which(raw_data$species %in% c("jackal", "lion"))] <- "Wildlife"

table(raw_data$species, raw_data$species_type)
```


<!----------------------------------------------------------------------------->
## Data preparation

Subsetting to only include 2014

```{r, echo = TRUE}
raw_data$date[1]
raw_data$date[1] %>% substr(1, 4)

map_data <- raw_data %>% 
  dplyr::mutate(year = substr(date, 1,4)) %>% 
  dplyr::filter(year == 2014)
```

<!----------------------------------------------------------------------------->
# Simple static maps

<!----------------------------------------------------------------------------->
## Map Basics

The simplest map of this data using base ``plot()``

```{r, echo = TRUE}
plot(map_data$y ~ map_data$x, asp = 1)
```

<!----------------------------------------------------------------------------->
## Map Basics

A simple verion using ``ggplot()`` with points coloured by human population density

```{r, echo = TRUE, fig.height = 4}
ggplot(data = map_data, aes(x = x, y = y, col = density)) +
  geom_point(alpha = 0.4) + # makes points transparent
  coord_equal() # sets aspect ratio 1:1
```

<!----------------------------------------------------------------------------->
## Map Basics {.smaller}

Combine previous plot with shape file data on regions

```{r, eval = F, echo = TRUE, fig.height = 3.5}
ggplot() +
  geom_polygon(data = region_shp_df, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = map_data, aes(x = x, y = y, col = density)) +
  coord_equal() # sets aspect ratio 1:1
```


<!----------------------------------------------------------------------------->
# Introduction to leaflet

<!----------------------------------------------------------------------------->
## Introduction to leaflet

### **What is leaflet?**

> - "Leaflet is a widely used open source JavaScript library used to build web mapping applications."

### **Why leaflet?**


> - Great way to explore datasets with geographic component
> - Communicate your results with interactive maps
> - Combine with shiny to produce even more interactive maps

<!----------------------------------------------------------------------------->
## Introduction to leaflet

### **What is leaflet?**
- open source JavaScript library used to build web mapping applications
- Probably the world's premier web mapping system

### **Why leaflet?**
- Great way to explore datasets with geographic component
- Communicate results with interactive maps using only a few lines of code
- Integrates nicely with Rshiny producing powerful extra-interactive maps

<!----------------------------------------------------------------------------->
## Intro to leaflet

```{r, eval = T, echo = T}
# insert fancy example leaflet map with several components
```

<!----------------------------------------------------------------------------->
## Getting started with our map {.smaller}

``leaflet()`` initialises a leaflet object ``m``. Pipe this into ``addCircles()`` where we refer to the ``map_data`` dataframe and indicate which columns correspond to latitude and longitude.

```{r, eval = T, echo = T, fig.height = 3.5}
m <- leaflet() %>%
  addCircles(data = map_data, lng = ~x, lat = ~y)
m 
```

<!----------------------------------------------------------------------------->
## Tiles {.smaller}

Tile give our data geographic context. ``addTiles()`` uses OpenStreetMap tiles, the default for leaflet

```{r, eval = T, echo = T, fig.height = 3.5}
m %>% addTiles()
```

<!----------------------------------------------------------------------------->
## Using ``addProviderTiles()`` {.smaller}

<b>Allows us to use third party tiles (e.g. Terrain tiles from Stamen Design).</b>

```{r, eval = T, echo = T}
# m created using code from previous slides
m <- leaflet() %>%
  addCircles(data=map_data, lng=~x, lat=~y)
```

<b>Code options</b>
```{r, eval = F, echo = T}
addProviderTiles(map = m, provider = "Stamen.Terrain")

addProviderTiles(map = m, provider = providers$Stamen.Terrain)

m %>% addProviderTiles(provider = providers$Stamen.Terrain)
```

<!----------------------------------------------------------------------------->
## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Stamen.Terrain)
```

<!----------------------------------------------------------------------------->
## Your turn

Try out a couple of the options for provider tiles.

Hint: Use tab completion with ``addProviderTiles(provideer = providers$)`` to access options

<!----------------------------------------------------------------------------->
## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.NatGeoWorldMap)
```

## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Wikimedia)
```

## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.WorldTopoMap)
```

## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.WorldImagery)
```

## Using ``addProviderTiles()`` {.smaller}
```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$CartoDB.Positron)
```


<!----------------------------------------------------------------------------->
## More info on  ``addProviderTiles()``

https://github.com/leaflet-extras/leaflet-providers
<br><br><br>
<b>Full list of free-to-use tiles with previews:</b> <br />
http://leaflet-extras.github.io/leaflet-providers/preview/index.html
<br><br><br>
<b>Tile servers based on OpenStreetMap data with info on how to cite:</b> <br />
https://wiki.openstreetmap.org/wiki/Tile_servers
<br><br><br>

e.g. When using the tiles provider$CartoDB.Positron:<br>
*Figure 1. Map of Tanzanian dataset. Cases coloured blue. Map tiles by Carto, under CC BY 3.0. Data by OpenStreetMap, under ODbL.*

<!----------------------------------------------------------------------------->
## Popups

**An interactive way for the user to explore a dataset**

Easier to first store what we want to display as an object ``popupInfo``
```{r, eval = T, echo = T}
popupInfo <- paste("Date: ", map_data$date, "<br>",
                   "Species: ", map_data$species, "<br>",
                   "Age: ", map_data$age, "<br>",
                   sep = " ")
```

Then, refer to ``popupInfo`` when using ``addCircles()``, ``addCircleMarkers()`` etc.

```{r, eval = T, echo = T}
m <- leaflet() %>% addTiles()
```


<!----------------------------------------------------------------------------->
## Popups

**An interactive way for the user to explore a dataset**

```{r, eval = T, echo = F}
m %>% addCircles(data = map_data, lng = ~x, lat = ~y, popup = popupInfo)
```

<!----------------------------------------------------------------------------->
## Marker types {.smaller}

Can use ``ifelse()`` to determine colour according to a variable.

```{r, eval = T, echo = T, fig.height=3.5}
m %>% addCircles(data=map_data, lng=~x, lat=~y, popup = popupInfo, stroke = T,
                 color = ifelse(map_data$species == "human", "firebrick", "dodgerblue"),
                 opacity = 0.8)
```


<!----------------------------------------------------------------------------->
# Colour palettes in leaflet

<!----------------------------------------------------------------------------->
## Discrete variable - using ``colorFactor()`` {.smaller}

``leaflet::colourFactor()`` creates a function (we're going to call it ``myPal``) that is used as an argument when we use ``addCircles()``, ``addCircleMarkers()``etc.

```{r, eval = T, echo = T}
# Choose 3 colours for our 3 groups  
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(palette = colour_pal,
                     domain = c("Human", "Domestic", "Wildlife"))
```

Create a fresh version of ``m`` before we add coloured circles

```{r, eval = T, echo = T}
m <- leaflet() %>% 
  addProviderTiles(provider = providers$CartoDB.Positron)
```

<!----------------------------------------------------------------------------->
## Discrete variable - using ``colorFactor()`` {.smaller}

``myPal()`` takes species type variable as argument

```{r, eval = T, echo = T, fig.height = 3.5}
m <- m %>% addCircles(data=map_data, lng=~x, lat=~y, 
                      color = myPal(map_data$species_type),
                      opacity = 0.9, popup = popupInfo)
m
```

<!----------------------------------------------------------------------------->
## Adding a legend

```{r, eval = T, echo = T, fig.height = 4}
m <-  m %>% addLegend(position = "topright", title = "Species<br>type", pal = myPal,
                      values = map_data$species_type, opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Your turn {.smaller}

Adapt the code from previous slides to produce a map with:<br>1) tiles showing topography<br>2) circles coloured by species (instead of species type)<br>3) a colour legend<br>4) a scale bar showing distance in kilometers (hint: ?addScaleBar).

Previous code:
```{r, eval = F, echo = T}
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(colour_pal, domain = c("Human", "Domestic", "Wildlife"))
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y,
             color = myPal(map_data$species_type), opacity = 0.9) %>%
  addLegend(position = "topright", title = "Species<br>type", pal = myPal,
            values = map_data$species_type, opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 1) tiles showing topography
```{r, eval = T, echo = T, fig.height = 3.5}
m <- leaflet() %>%
  addProviderTiles(provider = providers$OpenTopoMap)
m
  
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 2) circles coloured by species (instead of species type)
```{r, eval = T, echo = T, fig.height = 3.5}
colour_pal <- c("gold", "forestgreen", "cornflowerblue", "red", "magenta")
myPal <- colorFactor(colour_pal, domain = unique(map_data$species))
m <- m %>% addCircles(data = map_data, lng = ~x, lat = ~y,
                      color = myPal(map_data$species), opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 3) a colour legend

```{r, eval = T, echo = T, fig.height= 3.5}
m <- m %>% addLegend(position = "topright", title = "Species", pal = myPal,
                     values = map_data$species, opacity = 0.9)
m
```


<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 4) a scale bar showing distance in kilometers

```{r, eval = T, echo = T, fig.height= 3.5}
m <- m %>% addScaleBar(position = "bottomleft",
                       options = scaleBarOptions(maxWidth = 200, # default = 100
                                                 imperial = FALSE))
m
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date

First, need to convert date variable to a numeric value
```{r, eval = T, echo = T, fig.height = 4}
head(map_data$date, 5)
is.factor(map_data$date)

# Use 2 functions from lubridate for conversion
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date

First, need to convert date variable to a numeric value
```{r, eval = T, echo = T, fig.height = 4}
# Use 2 functions from lubridate for conversion
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))

head(map_data$date_decimal)
is.numeric(map_data$date_decimal)
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date

Provide some colours and values of date to ``colorNumeric()``.

Either provide colours (``palette = c("darkgreen", "green", "gold""))``) or refer to a palette (``palette = "Spectral"``). 

```{r, eval = T, echo = T, fig.height = 4}
dateRange <- c(2014, 2015)
myPal <- leaflet::colorNumeric(palette = "Spectral", domain = dateRange)

# new leaflet widget
m <- leaflet() %>% addProviderTiles(providers$CartoDB.Positron)
```
<!----------------------------------------------------------------------------->
## Continuous variable - colour by date {.smaller}

Add circles and a legend referring to ``myPal``

```{r, eval = T, echo = T, fig.height = 3.5}
m <- m %>% addCircles(data = map_data, lng = ~x, lat = ~y,
                      color = myPal(map_data$date_decimal), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Date",
                     pal = myPal, values = dateRange, opacity = 0.9,
                     labFormat = labelFormat(big.mark = "")) # removes a comma from 2,014
m
```


<!----------------------------------------------------------------------------->
## Your turn {.smaller}

Adapt the code from previous slides to produce a map with:
<br>1) Circles coloured by population density using the viridis palette
<br>2) and sized by population density (higher density = larger circles)
<br>3) legend and scale bar

Previous code to adapt:
```{r, eval = F, echo = T}
dateRange <- c(2014, 2015)
myPal <- colorNumeric(palette = "Spectral", domain = dateRange)
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y,
             color = myPal(map_data$date_decimal), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Date",
            pal = myPal, values = dateRange, opacity = 0.9,
            labFormat = labelFormat(big.mark = ""))
m
```

<!----------------------------------------------------------------------------->
## Solution

```{r, eval = T, echo = T, fig.height = 4}
densityRange <- c(0.1, 12000)
myPal <- colorNumeric(palette = "viridis", domain = log(densityRange))

m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y,
             color = myPal(log(map_data$density)), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Human<br>density<br>(log)",
            pal = myPal, values = log(densityRange), opacity = 0.9,
            labFormat = labelFormat(big.mark = "")) %>% 
  addScaleBar(position = "bottomleft")
m

```

## Continuous variable, discrete colour scale {.smaller}

```{r, eval = T, echo = T, fig.height = 4}
myPal <- colorQuantile(palette = "viridis", domain = map_data$density)

leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y, radius = 10,
             color = myPal(map_data$density), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Human<br>density",
            pal = myPal, values = map_data$density, opacity = 0.9,
            labFormat = labelFormat(big.mark = ""))

```



