---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "**Tanzania**, August, 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(leaflet)
library(rgdal)
raw_data <- read.csv("../../Day 1/data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- readOGR("../../Day 1/data/TZ_Region_2012", "TZ_Region_2012")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D") # 2:"#3E507F",
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
.blue {color: #4E23B8;}
.hl1-background {background-color: #fcf0cd;}
.hl2-background {background-color: #dfecfc;}

</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

  $('footnote').each(function(index) {
    var text  = $(this).html();
    var fnNum = (index+1).toString();
    $(this).html(fnNum.sup());

    var footnote   = fnNum + '. ' + text + '<br/>';
    var oldContent = $(this).parents('slide').children('div.footnotes').html();
    var newContent = oldContent + footnote;
    $(this).parents('slide').children('div.footnotes').html(newContent);
  });
});
</script>


<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Introduction to leaflet </b></li>
> - <b class="blue"> Building a leaflet map in R </b></li>
> - Building an interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!

<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

### **Day 1:**
> - Getting to know your data
> - Data subsetting, summarising
> - Building exploratory plots
> - Building an interactive plot in RShiny

### **Day 2:**
> - <b class="blue"> Introduction to leaflet </b></li>
> - <b class="blue"> Building *an interactive* leaflet map in R </b></li>
> - Building an **even more** interactive map in RShiny

### **Day 3:**
> - Review
> - Build your own apps!

<!----------------------------------------------------------------------------->
## Day 2

<ul>
<li> <b class="blue">1.1 Introduction to leaflet </b></li>
<li> 1.2 Building a leaflet map in R </li>
<li> 1.3 Building an interactive map in RShiny </li>
</ul>

<!----------------------------------------------------------------------------->
## Day 2

<ul>
<li> <b class="blue">1.1 Introduction to leaflet </b></li>
<li> 1.2 Building an interactive leaflet map in R </li>
<li> 1.3 Building an *even more* interactive map in RShiny </li>
</ul>

<!----------------------------------------------------------------------------->
## Recap day 1 {.smaller}

Remind ourselves what was in the data

```{r, echo = T}
# Set your working directory by clicking on the top menu:
# Session > Set Working Directory > To Source File Location
raw_data <- read.csv("data/raw_data.csv")
```

<!----------------------------------------------------------------------------->
## Recap day 1 {.smaller}

Reading in data from shape file

```{r, echo = T}
region_shp <- rgdal::readOGR("data/TZ_Region_2012", "TZ_Region_2012")
region_shp$id <- rownames(region_shp@data)
region_shp_df <- fortify(region_shp, region = "id")
region_shp_df <- left_join(region_shp_df, region_shp@data, by = "id")
```

<!----------------------------------------------------------------------------->
## Your Turn

> - Navigate to Day 2/worksheets, and open <b>1.0_worksheet.html</b>

> - All the activities we'll guide you to are in this file.

> - The contents table is interactive - you can click to move to the activity.

<br>

Complete section 2.1a of the worksheet

<!----------------------------------------------------------------------------->
## Map Basics {.smaller}

<b class="blue">How many entries does the data frame have?</b>

```{r, echo = TRUE}
# create variable for type of species
raw_data$species_type[which(raw_data$species=="dog" | raw_data$species=="cat")] <- "Domestic"
raw_data$species_type[which(raw_data$species %in% c("jackal", "lion"))] <- "Wildlife"
raw_data$species_type[which(raw_data$species=="human")] <- "Human"

# Use only one year
map_data <- raw_data %>% 
  dplyr::mutate(year = substr(date, 1,4)) %>% 
  dplyr::filter(year == 2014)
```

<!----------------------------------------------------------------------------->
## Map Basics {.smaller}

The simplest map of this data using base ``plot()``

```{r, echo = TRUE}
plot(map_data$y ~ map_data$x, asp = 1)
```

<!----------------------------------------------------------------------------->
## Map Basics {.smaller}

A simple verion using ``ggplot()`` with points coloured by human population density

```{r, echo = TRUE, fig.height = 4}
ggplot(data = map_data, aes(x = x, y = y, col = density)) +
  geom_point(alpha = 0.4) + # makes points transparent
  coord_equal() # sets aspect ratio 1:1
```

<!----------------------------------------------------------------------------->
## Map Basics {.smaller}

Combine previous plot with shape file data on regions

```{r, eval = F, echo = TRUE, fig.height = 3.5}
ggplot() +
  geom_polygon(data = region_shp_df, aes(x = long, y = lat, group = group), 
               fill = "white", color = "black") +
  geom_point(data = map_data, aes(x = x, y = y, col = density)) +
  coord_equal() # sets aspect ratio 1:1
```


<!----------------------------------------------------------------------------->
# Introduction to leaflet

<!----------------------------------------------------------------------------->
## Introduction to leaflet

### **What is leaflet?**

> - "Leaflet is a widely used open source JavaScript library used to build web mapping applications."

### **Why leaflet?**


> - Great way to explore datasets with geographic component
> - Communicate your results with interactive maps
> - Combine with shiny to produce even more interactive maps

<!----------------------------------------------------------------------------->
## Introduction to leaflet

### **What is leaflet?**
- open source JavaScript library used to build web mapping applications
- Probably the world's premier web mapping system

### **Why leaflet?**
- Great way to explore datasets with geographic component
- Communicate results with interactive maps using only a few lines of code
- Integrates nicely with Rshiny producing powerful extra-interactive maps

<!----------------------------------------------------------------------------->
## Introduction to leaflet

```{r }
# insert fancy example leaflet map with several components
```

<!----------------------------------------------------------------------------->
## Getting started with our map {.smaller}

<b>We'll create a leaflet object ``m`` and add features, one step at a time.</b>

<!---setView(lng = 38.9, lat = -6.4333, zoom = 12)
m %>% addTiles()--->

We'll always start by using ``leaflet()`` to create a default widget
```{r, eval = T, echo = T}
m <- leaflet()
```

Next, add circles for our data points using ``addCircles()``
```{r, eval = T, echo = T}
m <- m %>%
  addCircles(data=map_data, lng=~x, lat=~y)

m <- addCircles(m, data=map_data, lng=~x, lat=~y)
m
```

<!----------------------------------------------------------------------------->
## Using tiles to put data in geographic context {.smaller}

``addTiles()`` uses OpenStreetMap tiles, the default for leaflet

```{r, eval = T, echo = T}
# use the pipe operator to feed m into addTiles():
m %>% addTiles()
```

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}

<b>Allows us to use third party tiles (e.g. Terrain tiles from Stamen Design).</b>

```{r, eval = T, echo = T}
# m created using code from previous slides
m <- leaflet() %>%
  addCircles(data=map_data, lng=~x, lat=~y)
```

<b>Code options</b>
```{r, eval = F, echo = T}
addProviderTiles(map = m, provider = "Stamen.Terrain")

addProviderTiles(map = m, provider = providers$Stamen.Terrain)

m %>% addProviderTiles(provider = providers$Stamen.Terrain)
```

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}
```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Stamen.Terrain)
```

<!----------------------------------------------------------------------------->
## Your turn

Try out a couple of the options for provider tiles.

Hint: Use tab completion with ``addProviderTiles(providers$)`` to access options

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.NatGeoWorldMap)
```

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}
```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Wikimedia)
```

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.WorldTopoMap)
```

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}
```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.WorldImagery)
```

<!----------------------------------------------------------------------------->
## Alternative tiles using ``addProviderTiles()`` {.smaller}
```{r, eval = T, echo = T}
m <- m %>% addProviderTiles(providers$CartoDB.Positron)
m
```


<!----------------------------------------------------------------------------->
## More info on  ``addProviderTiles()`` {.smaller}

https://github.com/leaflet-extras/leaflet-providers
<br><br><br>
<b>Full list of free-to-use tiles with previews:</b> <br />
http://leaflet-extras.github.io/leaflet-providers/preview/index.html
<br><br><br>
<b>Tile servers based on OpenStreetMap data with info on how to cite:</b> <br />
https://wiki.openstreetmap.org/wiki/Tile_servers
<br><br><br>

e.g. When using the tiles provider$CartoDB.Positron:<br>
*Figure 1. Map of Tanzanian dataset. Cases coloured blue. Map tiles by Carto, under CC BY 3.0. Data by OpenStreetMap, under ODbL.*

<!----------------------------------------------------------------------------->
## Editing data icons {.smaller}

```{r, eval = T, echo = T}
?addCircles
```

<!----------------------------------------------------------------------------->
## Popups {.smaller}

Interactive way for user to explore dataset

```{r, eval = T, echo = T}
# We can create an object which has everything we want to display
popupInfo <- paste("Date: ", map_data$date, "<br>",
                   "Species: ", map_data$species, "<br>",
                   "Age: ", map_data$age, "<br>",
                   sep = " ")
```

Then, refer to ``popupInfo`` when using ``addCircles()``, ``addCircleMarkers()`` etc.

```{r, eval = T, echo = T}
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(data=map_data, lng=~x, lat=~y, popup = popupInfo)
  
```

<!----------------------------------------------------------------------------->
## Popups {.smaller}

**An interactive way for the user to explore a dataset**

Easier to first store what we want to display as an object ``popupInfo``
```{r, eval = T, echo = T}
popupInfo <- paste("Date: ", map_data$date, "<br>",
                   "Species: ", map_data$species, "<br>",
                   "Age: ", map_data$age, "<br>",
                   sep = " ")
```

Then, refer to ``popupInfo`` when using ``addCircles()``, ``addCircleMarkers()`` etc.

```{r, eval = T, echo = T}
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(data=map_data, lng=~x, lat=~y, popup = popupInfo)
```


<!----------------------------------------------------------------------------->
## Popups {.smaller}

**An interactive way for the user to explore a dataset**

```{r, eval = T, echo = T}
m
```

<!----------------------------------------------------------------------------->
## Introducing colour palettes in leaflet

**Using the ``colorFactor()`` and ``colorNumeric()`` functions**

<!----------------------------------------------------------------------------->
## Discrete variable - colour by species type {.smaller}

Create a fresh version of ``m``:
```{r, eval = T, echo = T}
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron)
```

``leaflet::colourFactor()`` creates a function (we're going to call it ``myPal``) that is used as an argument when we use ``addCircles()``

```{r, eval = T, echo = T}
# Choose 3 colours for our 3 groups  
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(colour_pal,
                     domain = c("Human", "Domestic", "Wildlife"))
```

<!----------------------------------------------------------------------------->
## Discrete variable - colour by species type {.smaller}

```{r, eval = T, echo = T, fig.height = 4}
# myPal() takes species type variable as argument
m <- m %>% addCircles(data=map_data, lng=~x, lat=~y,
                      color = myPal(map_data$species_type), opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Adding a legend {.smaller}

```{r, eval = T, echo = T, fig.height = 4}
m <-  m %>% addLegend(position = "topright", title = "Species<br>type", pal = myPal,
                      values = map_data$species_type, opacity = 0.5)
m
```

<!----------------------------------------------------------------------------->
## Your turn {.smaller}

Adapt the code from previous slides to produce a map with:<br>1) tiles showing elevation<br>2) circles coloured by species (instead of species type)<br>3) a legend for colour<br>4) a scale bar showing distance in kilometers (hint: ?addScaleBar).

Previous code:
```{r, eval = F, echo = T}
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(colour_pal, domain = c("Human", "Domestic", "Wildlife"))
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircles(data=map_data, lng=~x, lat=~y,
             color = myPal(map_data$species_type), opacity = 0.9) %>%
  addLegend(position = "topright", title = "Species<br>type", pal = myPal,
            values = map_data$species_type, opacity = 0.5)
m
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 1) tiles showing elevation
```{r, eval = T, echo = T}
m <- leaflet() %>%
  addProviderTiles(providers$OpenTopoMap)
m
  
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 2) circles coloured by species (instead of species type)
```{r, eval = T, echo = T}
colour_pal <- c("gold", "forestgreen", "cornflowerblue", "red", "magenta")
myPal <- colorFactor(colour_pal, domain = unique(map_data$species))
m <- m %>% addCircles(data = map_data, lng = ~x, lat = ~y,
                      color = myPal(map_data$species), opacity = 0.9)
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 3) a legend for colour

```{r, eval = T, echo = T}
m <- m %>% addLegend(position = "topright", title = "Species", pal = myPal,
                     values = map_data$species, opacity = 0.5)
```


<!----------------------------------------------------------------------------->
## Solution {.smaller}

Part 4) a scale bar showing distance in kilometers

```{r, eval = T, echo = T, fig.height= 3.2}
m <- m %>% addScaleBar(position = "bottomleft",
                       options = scaleBarOptions(maxWidth = 200, # default = 100
                                                 imperial = FALSE))
m
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

```{r, eval = T}
m
```


<!----------------------------------------------------------------------------->
## Continuous variable - colour by date {.smaller}

First, need to convert date variable to a numeric value
```{r, eval = T, echo = T, fig.height = 4}
head(map_data$date, 5)
is.character(map_data$date)

# Use 2 functions from lubridate for conversion
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date {.smaller}

First, need to convert date variable to a numeric value
```{r, eval = T, echo = T, fig.height = 4}
# Use 2 functions from lubridate for conversion
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))

head(map_data$date_decimal)
is.numeric(map_data$date_decimal)
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date {.smaller}

```{r, eval = T, echo = T, fig.height = 4}
dateRange <- c(2014, 2015)
myPal <- leaflet::colorNumeric(col_palette, dateRange)
m <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircles(data=map_data, lng=~x, lat=~y,
             color = myPal(map_data$date_decimal), opacity = 0.9)
```

Add a legend. ``labFormat = labelFormat(big.mark = "")`` removes a comma from 2,014

```{r, eval = T, echo = T, fig.height = 4}  
m <- m %>% addLegend(position = "topright", title = "Date",
                     pal = myPal, values = dateRange,
                     labFormat = labelFormat(big.mark = ""))
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date {.smaller}

```{r, eval = T, echo = T, fig.height = 4}
m %>% addScaleBar(position = "bottomleft")
```
