---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "**Tanzania**, August, 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(leaflet)
library(rgdal)
raw_data <- read.csv("../../Day 1/data/raw_data.csv", stringsAsFactors = FALSE)
raw_data$species_type <- ifelse(raw_data$species == "human", "Human", NA)
raw_data$species_type[which(raw_data$species %in% c("dog", "cat"))] <- "Domestic"
raw_data$species_type[which(raw_data$species %in% c("jackal", "lion"))] <- "Wildlife"
map_data <- raw_data %>% 
  dplyr::mutate(year = substr(date, 1,4)) %>% 
  dplyr::filter(year == 2014)
# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D") # 2:"#3E507F",

# extract mean density from data in each region
mean_density <- data.frame(tapply(X = raw_data$density,
                                  INDEX = list(raw_data$region),
                                  FUN = mean))
names(mean_density) <- c("density")
mean_density$region <- row.names(mean_density)
head(mean_density)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
.blue {color: #4E23B8;}
.hl1-background {background-color: #fcf0cd;}
.hl2-background {background-color: #dfecfc;}

</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

  $('footnote').each(function(index) {
    var text  = $(this).html();
    var fnNum = (index+1).toString();
    $(this).html(fnNum.sup());

    var footnote   = fnNum + '. ' + text + '<br/>';
    var oldContent = $(this).parents('slide').children('div.footnotes').html();
    var newContent = oldContent + footnote;
    $(this).parents('slide').children('div.footnotes').html(newContent);
  });
});
</script>

<!----------------------------------------------------------------------------->
# Using leaflet and shape file data

<!----------------------------------------------------------------------------->
## Shape file data {.smaller}
 
```{r, eval = T, echo = T, message = F, fig.height = 4}
region_shp <- rgdal::readOGR(dsn = "data/TZ_Region_2012_small")
names(region_shp)
```

<!----------------------------------------------------------------------------->
## Shape file data {.smaller}

Use ``addPolygons()`` to add to leaflet map. ``label = `` determines what will be shown when cursor hovers above area.

```{r, eval = T, echo = T, message = F, fig.height = 3.5}
leaflet() %>% addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>% 
  addPolygons(data = region_shp, label = region_shp$Region_Nam,
              color = "black", weight = 1, fillOpacity = 0.3, opacity = 0.6)
```

<!----------------------------------------------------------------------------->
## Shape file data {.smaller}

Use ``highlightOptions()`` to highlight shapes when cursor hovers above. ``bringToFront = TRUE`` avoids potential problems with order in which shapes are coded in shape file.

```{r, eval = T, echo = T, message = F, fig.height = 3.5}
leaflet() %>% addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>% 
  addPolygons(data = region_shp, label = region_shp$Region_Nam, fillColor = "gold",
              color = "black", weight = 1, fillOpacity = 0.3, opacity = 0.6,
              highlightOptions = highlightOptions(color = "white", weight = 3,
                                                  bringToFront = TRUE, opacity = 1))
```

<!----------------------------------------------------------------------------->
## Shape file data {.smaller}

Possible to use ``addPolylines()`` to add outlines only to leaflet map.

```{r, eval = T, echo = T, message = F, fig.height = 4}
leaflet() %>% addProviderTiles(provider = providers$Esri.WorldShadedRelief) %>% 
  addPolylines(data = region_shp, color = "red",  weight = 2)
```

<!----------------------------------------------------------------------------->
## Your turn

Produce a leaflet map using ``addPolygons()`` and ``addPolylines()`` with tiles and circle markers coloured by human population density, a colour legend and a scale bar.

```{r, eval = T, echo = T, message = F, fig.height = 4}
colour_pal <- RColorBrewer::brewer.pal(11, "Spectral")
range(map_data$density)
myPal <- leaflet::colorNumeric(palette = colour_pal, domain = log(map_data$density))

region_shp$density <- NA
for (i in 1:length(region_shp$density)) {
  region_shp$density[i] <- mean_density$density[mean_density$region == region_shp$Region_Nam[i]]
}

labels <- sprintf("<strong>%s</strong><br/>%g people / mi<sup>2</sup>",
                  region_shp$Region_Nam, region_shp$density) %>% lapply(htmltools::HTML)

pal <- colorQuantile("YlOrRd", domain = region_shp$density)
m %>% addPolygons(data = region_shp, fillColor = pal(region_shp$density),
                  label = labels, weight = 2, opacity = 1, color = "white",
                  fillOpacity = 0.7)
```

<!----------------------------------------------------------------------------->
## Solution {.smaller}

```{r, eval = T, echo = T, message = F, fig.height = 4}
leaflet() %>%
  addProviderTiles(provider = providers$Esri.WorldTopoMap) %>%
  addPolylines(data = region_shp, color = "black",  weight = 2) %>% 
  addPolygons(data = region_shp, label = region_shp$Region_Nam,
              color = "grey", weight = 1, fillOpacity = 0.3,
              highlightOptions = highlightOptions(color = "white", weight = 2,
                                                  bringToFront = F)) %>%
  addCircleMarkers(data = map_data, lng = ~x, lat = ~y,
                   radius = 5, fillColor = myPal(log(map_data$density)), fillOpacity = 0.8,
                   weight = 2, color = "grey", opacity = 0.9) %>% 
  addLegend(pal = myPal, values = log(map_data$density), title = "Human<br>population<br>density")
```

<!----------------------------------------------------------------------------->
## Chloropleth

Colour shapes for each region by some value. First we add a value for each region to the shape file object.

```{r, eval = T, echo = T, fig.height = 4}
head(mean_density)

```

<!----------------------------------------------------------------------------->
# Using leaflet and environmental data in raster format

<!----------------------------------------------------------------------------->
## Loading raster data {.smaller}

Load data using raster() from the raster package

```{r, eval = T, echo = T, fig.height = 4}
density <- raster::raster("../../Day 1/data/HumanPopulation.grd")
density
```

<!----------------------------------------------------------------------------->
## Plotting raster data raster data {.smaller}

Load data using raster() from the raster package

```{r, eval = T, echo = T, fig.height = 4}
plot(density)
```

<!----------------------------------------------------------------------------->
## Plotting raster data {.smaller}

Load data using raster() from the raster package

```{r, eval = T, echo = T, fig.height = 4}
plot(log(density))
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

We use ``colorNumeric()`` again, as we did to colour by date earlier

```{r, eval = T, echo = T, fig.height = 4}
colour_pal <- c("#FFFFCC", "#41B6C4", "#0C2C84")
raster::values(density) %>% range(na.rm = T)
myPal <- colorNumeric(palette = colour_pal, domain = c(0, 30000), na.color = "red")

m <- leaflet() %>%
  addRasterImage(x = density, colors = myPal, opacity = 0.8)
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

```{r, eval = T, echo = F, fig.height = 4}
m
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

Let's log density as we did before

```{r, eval = T, echo = T, fig.height = 3.5}
raster::values(density) %>% range(na.rm = T) %>% log
myPal <- colorNumeric(palette = colour_pal, domain = c(-1.5, 11),
                      na.color = "transparent")
```

We have to exclude some data, as there are 0 density polygons and log(0) gives -infinity

```{r, eval = T, echo = T, fig.height = 3.5}
sum(raster::values(density) > exp(-1.5), na.rm = T) /
  sum(raster::values(density) >= 0, na.rm = T)
```


<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}


Plot the logged raster data with some tiles and a legend for the colour scheme.

```{r, eval = T, echo = T, fig.height = 3.5}
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addRasterImage(x = log(density), colors = myPal, opacity = 0.8) %>%
  addLegend(pal = myPal, values = c(-1.5, 11), opacity = 0.8,
            title = "Human<br>population<br>density")
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

```{r, eval = T, echo = F, fig.height = 4}
m
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}


Instead of using ``addProviderTiles(provider = providers$CartoDB.Positron)``

We order things here so that we plot<br>1) The background tiles (``providers$CartoDB.PositronNoLabels``) only first<br>2) Then the raster data<br>3) and finally the labels (``providers$CartoDB.PositronOnlyLabels``) on top of that

```{r, eval = T, echo = T, fig.height = 3.5}
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.PositronNoLabels) %>%
  addRasterImage(x = log(density), colors = myPal, opacity = 0.8) %>%
  addProviderTiles(provider = providers$CartoDB.PositronOnlyLabels) %>%
  addLegend(pal = myPal, values = c(-1.5, 11), opacity = 0.8,
            title = "Human<br>population<br>density")
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

```{r, eval = T, echo = F, fig.height = 4}
m
```
