---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "**Tanzania**, August, 2019"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(leaflet)
library(rgdal)
library(mapview)
library(magrittr)
raw_data <- read.csv("../../Day 1/data/raw_data.csv", stringsAsFactors = FALSE)
raw_data$species_type <- ifelse(raw_data$species == "human", "Human", NA)
raw_data$species_type[which(raw_data$species %in% c("dog", "cat"))] <- "Domestic"
raw_data$species_type[which(raw_data$species %in% c("jackal", "lion"))] <- "Wildlife"
map_data <- raw_data %>% 
  dplyr::mutate(year = substr(date, 1,4)) %>% 
  dplyr::filter(year == 2014)
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3.5)
```

<style>
div.footnotes {
  position: absolute;
  bottom: 0;
  margin-bottom: 10px;
  width: 80%;
  font-size: 0.6em;
}
.blue {color: #4E23B8;}
.hl1-background {background-color: #fcf0cd;}
.hl2-background {background-color: #dfecfc;}

</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
  $('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

  $('footnote').each(function(index) {
    var text  = $(this).html();
    var fnNum = (index+1).toString();
    $(this).html(fnNum.sup());

    var footnote   = fnNum + '. ' + text + '<br/>';
    var oldContent = $(this).parents('slide').children('div.footnotes').html();
    var newContent = oldContent + footnote;
    $(this).parents('slide').children('div.footnotes').html(newContent);
  });
});
</script>

<!----------------------------------------------------------------------------->
# Introduction to leaflet

<!----------------------------------------------------------------------------->
## Introduction to leaflet

### **What is leaflet?**
- open source JavaScript library used to build web mapping applications
- Probably the world's premier web mapping system

### **Why leaflet?**
- Great way to explore datasets with geographic component
- Communicate results with interactive maps using only a few lines of code
- Integrates nicely with Rshiny producing powerful extra-interactive maps

<!----------------------------------------------------------------------------->
## Intro to leaflet

```{r, eval = T, echo = T}
# insert fancy example leaflet map with several components
1 + 1
```

<!----------------------------------------------------------------------------->
## Getting started with our map {.smaller}

``leaflet()`` initialises a leaflet object ``m``. Pipe this into ``addCircles()`` where we refer to the ``map_data`` dataframe and indicate which columns correspond to latitude and longitude.

```{r, eval = T, echo = T, fig.height = 3.5}
m <- leaflet() %>%
  addCircles(data = map_data, lng = ~x, lat = ~y)
m 
```

<!----------------------------------------------------------------------------->
## Tiles {.smaller}

Tile give our data geographic context. ``addTiles()`` uses OpenStreetMap tiles, the default for leaflet

```{r, eval = T, echo = T, fig.height = 3.5}
m %>% addTiles()
```

<!----------------------------------------------------------------------------->
## Using ``addProviderTiles()`` {.smaller}

<b>Allows us to use third party tiles (e.g. Terrain tiles from Stamen Design).</b>

```{r, eval = T, echo = T}
# m created using code from previous slides
m <- leaflet() %>%
  addCircles(data=map_data, lng=~x, lat=~y)
```

<b>Code options</b>
```{r, eval = F, echo = T}
addProviderTiles(map = m, provider = "Stamen.Terrain")

addProviderTiles(map = m, provider = providers$Stamen.Terrain)

m %>% addProviderTiles(provider = providers$Stamen.Terrain)
```

<!----------------------------------------------------------------------------->
## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Stamen.Terrain)
```

<!----------------------------------------------------------------------------->
## Your turn 2.2a

Try some options for provider tiles.

Hint: Use tab completion with ``addProviderTiles(provideer = providers$)`` to access options

<!----------------------------------------------------------------------------->
## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.NatGeoWorldMap)
```

## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Wikimedia)
```

## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.WorldTopoMap)
```

## Using ``addProviderTiles()`` {.smaller}

```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$Esri.WorldImagery)
```

## Using ``addProviderTiles()`` {.smaller}
```{r, eval = T, echo = T}
m %>% addProviderTiles(providers$CartoDB.Positron)
```


<!----------------------------------------------------------------------------->
## More info on  ``addProviderTiles()``

https://github.com/leaflet-extras/leaflet-providers
<br><br><br>
<b>Full list of free-to-use tiles with previews:</b> <br />
http://leaflet-extras.github.io/leaflet-providers/preview/index.html
<br><br><br>
<b>Tile servers based on OpenStreetMap data with info on how to cite:</b> <br />
https://wiki.openstreetmap.org/wiki/Tile_servers
<br><br><br>

e.g. When using the tiles provider$CartoDB.Positron:<br>
*Figure 1. Map of Tanzanian dataset. Cases coloured blue. Map tiles by Carto, under CC BY 3.0. Data by OpenStreetMap, under ODbL.*

<!----------------------------------------------------------------------------->
## Popups

**An interactive way for the user to get more info on data points**

Easier to first store what we want to display as an object ``popupInfo``
```{r, eval = T, echo = T}
popupInfo <- paste("Date: ", map_data$date, "<br>",
                   "Species: ", map_data$species, "<br>",
                   "Age: ", map_data$age, "<br>",
                   sep = " ")

m <- leaflet() %>% addProviderTiles(providers$CartoDB.Positron)
```

Then, refer to ``popupInfo`` when using ``addCircles()``, ``addCircleMarkers()`` etc.

```{r, eval=F}
m %>% addCircles(data = map_data, lng = ~x, lat = ~y, popup = popupInfo)
```


<!----------------------------------------------------------------------------->
## Popups

**An interactive way for the user to get more info on data points**

```{r}
m %>% addCircles(data = map_data, lng = ~x, lat = ~y, popup = popupInfo)
```

<!----------------------------------------------------------------------------->
## Popups and labels

Labels show up when hovered over, popups when the user clicks.

```{r}
m %>% addCircles(data = map_data, lng = ~x, lat = ~y, popup = popupInfo,
                 label = ~species)
```

<!----------------------------------------------------------------------------->
## Circle markers {.smaller}

Circle markers (``addCircleMarkers()``) are similar to circles but radius in onscreen pixels is constant regardless of zoom level.

```{r}
m %>% addCircleMarkers(data=map_data, lng=~x, lat=~y, popup = popupInfo, radius = 5,
                       stroke = T, weight = 3, color = "grey", opacity = 0.8,
                       fillColor = ifelse(map_data$species == "human",
                                          "firebrick", "dodgerblue"), fillOpacity = 0.7)
```

<!----------------------------------------------------------------------------->
## Circle markers {.smaller}

With, ``stroke = F``, weight, color, and opacity are irrelevant. Can also use ``radius = `` to size circle markers by species.
 
```{r}
m %>% addCircleMarkers(data = map_data, lng = ~x, lat = ~y, popup = popupInfo, stroke = F,
                       radius = ifelse(map_data$species == "human", 8, 4),
                       fillColor = ifelse(map_data$species == "human", "firebrick", "dodgerblue"),
                       fillOpacity = 0.8)
```

<!----------------------------------------------------------------------------->
## addMarkers {.smaller}

On a map with many markers, it might make sense to cluster them using ``clusterOptions = markerClusterOptions()`` shown here with the default options for ``markerClusterOptions()``.
 
```{r}
m %>% addMarkers(data = map_data, lng = ~x, lat = ~y, label = ~species, popup = popupInfo)
```

<!----------------------------------------------------------------------------->
## addMarkers {.smaller}

On a map with many markers, it might make sense to cluster them using ``clusterOptions = markerClusterOptions()`` shown here with the default options for ``markerClusterOptions()``.
 
```{r}
m %>% addMarkers(data = map_data, lng = ~x, lat = ~y, label = ~species, popup = popupInfo,
                 clusterOptions = markerClusterOptions())
```



<!----------------------------------------------------------------------------->
# Colour palettes in leaflet

<!----------------------------------------------------------------------------->
## Discrete variable - using ``colorFactor()`` {.smaller}

``leaflet::colourFactor()`` creates a function (we're going to call it ``myPal``) that is used as an argument when we use ``addCircles()``, ``addCircleMarkers()``etc.

```{r}
# Choose 3 colours for our 3 groups  
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(palette = colour_pal,
                     domain = c("Human", "Domestic", "Wildlife"))
```

Create a fresh version of ``m`` before we add coloured circles

```{r}
m <- leaflet() %>% 
  addProviderTiles(provider = providers$CartoDB.Positron)
```

<!----------------------------------------------------------------------------->
## Discrete variable - using ``colorFactor()`` {.smaller}

``myPal()`` takes species type variable as argument

```{r}
m <- m %>% addCircles(data = map_data, lng = ~x, lat = ~y, 
                      color = myPal(map_data$species_type),
                      opacity = 0.9, popup = popupInfo)
m
```

<!----------------------------------------------------------------------------->
## Adding a legend

```{r}
m <-  m %>% addLegend(position = "topright", title = "Species<br>type", pal = myPal,
                      values = map_data$species_type, opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Your turn 2.2b {.smaller}

Adapt the code from previous slides to produce a map with:<br>1) tiles showing topography<br>2) circles coloured by species (instead of species type)<br>3) a colour legend<br>4) a scale bar showing distance in kilometers (hint: ?addScaleBar).

Previous code:
```{r, eval=F}
colour_pal <- c("gold", "forestgreen", "cornflowerblue")
myPal <- colorFactor(colour_pal, domain = c("Human", "Domestic", "Wildlife"))
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y,
             color = myPal(map_data$species_type), opacity = 0.9) %>%
  addLegend(position = "topright", title = "Species<br>type", pal = myPal,
            values = map_data$species_type, opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Solution 2.2b {.smaller}

Part 1) tiles showing topography

```{r}
m <- leaflet() %>%
  addProviderTiles(provider = providers$Esri.WorldPhysical)
m %>% setView(lat = -6.4333, lng = 38.9, zoom = 5)
  
```

<!----------------------------------------------------------------------------->
## Solution 2.2b {.smaller}

Part 2) circles coloured by species (instead of species type)

```{r}
colour_pal <- c("gold", "forestgreen", "cornflowerblue", "red", "magenta")
myPal <- colorFactor(palette = colour_pal, domain = unique(map_data$species))
m <- m %>% addCircles(data = map_data, lng = ~x, lat = ~y,
                      color = myPal(map_data$species), opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Solution 2.2b {.smaller}

Part 2) circles coloured by species (instead of species type)

```{r}
myPal <- colorFactor(palette = "Set2", domain = unique(map_data$species))
m <- m %>% addCircles(data = map_data, lng = ~x, lat = ~y,
                      color = myPal(map_data$species), opacity = 0.9)
m
```

<!----------------------------------------------------------------------------->
## Solution 2.2b {.smaller}

Part 3) a colour legend

```{r, eval = T, echo = T, fig.height= 3.5}
m <- m %>% addLegend(position = "topright", title = "Species", pal = myPal,
                     values = map_data$species, opacity = 0.9)
m
```


<!----------------------------------------------------------------------------->
## Solution 2.2b {.smaller}

Part 4) a scale bar showing distance in kilometers

```{r}
m <- m %>% addScaleBar(position = "bottomleft",
                       options = scaleBarOptions(maxWidth = 200, # default = 100
                                                 imperial = FALSE))
m
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date

First, need to convert date variable to a numeric value

```{r}
head(map_data$date, 5)
is.factor(map_data$date)

# Use 2 functions from lubridate for conversion
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date

date_decimal is now a numeric variable

```{r}
# Use 2 functions from lubridate for conversion
map_data$date_decimal <- lubridate::decimal_date(lubridate::ymd(map_data$date))

head(map_data$date_decimal)
is.numeric(map_data$date_decimal)
```

<!----------------------------------------------------------------------------->
## Continuous variable - colour by date

Provide some colours and values of date to ``colorNumeric()``.

Either provide colours (``palette = c("darkgreen", "green", "gold""))``) or refer to a palette (``palette = "Spectral"``). 

```{r}
dateRange <- c(2014, 2015)
myPal <- leaflet::colorNumeric(palette = "Spectral", domain = dateRange)

# new leaflet widget
m <- leaflet() %>% addProviderTiles(providers$CartoDB.Positron)
```
<!----------------------------------------------------------------------------->
## Continuous variable - colour by date {.smaller}

Add circles and a legend referring to ``myPal``

```{r}
m %>% addCircles(data = map_data, lng = ~x, lat = ~y,
                 color = myPal(map_data$date_decimal), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Date",
            pal = myPal, values = dateRange, opacity = 0.9,
            labFormat = labelFormat(big.mark = "")) # removes a comma from 2,014
```


<!----------------------------------------------------------------------------->
## Your turn 2.2c {.smaller}

Adapt the code from previous slides to produce a map with:
<br>1) Circles coloured by population density using the viridis palette
<br>2) and sized by population density (higher density = larger circles)
<br>3) legend and scale bar

Previous code to adapt:
```{r, eval = F}
dateRange <- c(2014, 2015)
myPal <- colorNumeric(palette = "Spectral", domain = dateRange)
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y,
             color = myPal(map_data$date_decimal), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Date",
            pal = myPal, values = dateRange, opacity = 0.9,
            labFormat = labelFormat(big.mark = ""))
m
```

<!----------------------------------------------------------------------------->
## Solution 2.2c

```{r}
densityRange <- c(0.1, 12000)
myPal <- colorNumeric(palette = "viridis", domain = log(densityRange))

m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y,
             color = myPal(log(map_data$density)), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Human<br>density<br>(log)",
            pal = myPal, values = log(densityRange), opacity = 0.9) %>% 
  addScaleBar(position = "bottomleft")
```

<!----------------------------------------------------------------------------->
## Solution 2.2c

```{r}
m
```

## Continuous variable, discrete colour scale {.smaller}

```{r, fig.height=3.2}
myPal <- colorQuantile(palette = "viridis", domain = map_data$density)

leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addCircles(data = map_data, lng = ~x, lat = ~y, radius = 10,
             color = myPal(map_data$density), opacity = 0.9) %>% 
  addLegend(position = "topright", title = "Human<br>density",
            pal = myPal, values = map_data$density, opacity = 0.9,
            labFormat = labelFormat(big.mark = ""))

```

<!----------------------------------------------------------------------------->
# Using leaflet and shape file data

<!----------------------------------------------------------------------------->
## Shape file data {.smaller}

Load shape data and look at the structure

```{r, eval = T, message = F}
region_shp <- rgdal::readOGR(dsn = "data/TZ_Region_2012_density")
head(region_shp)
```

<!----------------------------------------------------------------------------->
## Plotting polygons {.smaller}

Use ``addPolygons()`` to add to leaflet map. ``label = `` determines what will be shown when cursor hovers above area.

```{r, eval = T, message = F}
m <- leaflet() %>% addProviderTiles(provider = providers$Esri.WorldShadedRelief)
m %>% addPolygons(data = region_shp, label = region_shp$Region_Nam,
                  color = "black", weight = 1, fillOpacity = 0.3, opacity = 0.6)
```

<!----------------------------------------------------------------------------->
## Plotting polygons {.smaller}

Use ``highlightOptions()`` to highlight shapes when cursor hovers above. ``bringToFront = TRUE`` avoids potential problems with order in which shapes are coded in shape file.

```{r, eval = T, message = F}
m %>% addPolygons(data = region_shp, label = region_shp$Region_Nam, fillColor = "gold",
                  color = "black", weight = 1, fillOpacity = 0.3, opacity = 0.6,
                  highlightOptions = highlightOptions(color = "white", weight = 3,
                                                      bringToFront = TRUE, opacity = 1))
```

<!----------------------------------------------------------------------------->
## Plotting polygons {.smaller}

Possible to use ``addPolylines()`` to add outlines only to leaflet map.

```{r, eval = T, message = F}
m %>% addPolylines(data = region_shp, color = "red",  weight = 2, opacity = 1)
```

<!----------------------------------------------------------------------------->
## Plotting a choropleth {.smaller}

Choropleth: Map with areas shaded according to some numeric variable
<br>(similar to heatmap but uses geographic boundaries)

```{r, eval = T}
myPal <- colorNumeric(palette = "YlOrRd", domain = region_shp$density)
m %>% addPolygons(data = region_shp, label = region_shp$Region_Nam,
                  fillColor = myPal(region_shp$density), fillOpacity = 0.7,
                  color = "white", opacity = 1, weight = 2)
```

<!----------------------------------------------------------------------------->
## Plotting a choropleth {.smaller}

Like before, we can log density to give a more informative map

```{r}
myPal <- colorNumeric(palette = "YlOrRd", domain = log(region_shp$density))
m %>% addPolygons(data = region_shp, label = region_shp$Region_Nam,
                  fillColor = myPal(log(region_shp$density)), fillOpacity = 0.7,
                  color = "white", opacity = 1, weight = 2)
```

<!----------------------------------------------------------------------------->
## Your turn 2.2d

Adjust the choropleth code below to:

1) use ``colorQuantile()`` instead of ``colorNumeric()``
2) add circles coloured by date for human cases

```{r eval = F}
m <- leaflet() %>% addProviderTiles(provider = providers$Esri.WorldShadedRelief)

myPal <- colorNumeric(palette = "YlOrRd", domain = log(region_shp$density))
m %>% addPolygons(data = region_shp, label = region_shp$density,
                  fillColor = myPal(log(region_shp$density)), fillOpacity = 0.7,
                  color = "white", opacity = 1, weight = 2)
```


<!----------------------------------------------------------------------------->
## Solution 2.2d

Subset using ``map_data[map_data$species == "human",]`` or alternatively ``dplyr::select()``

With ``colorQuantile()``, we don't need to log density

Create a second colour palette for data with ``colorNumeric()``.

```{r, eval = F}
human_data <- map_data[map_data$species == "human",]
myPal <- colorQuantile(palette = "YlOrRd", domain = region_shp$density)
myPal2 <- colorNumeric(palette = "viridis", domain = human_data$date_decimal)
m %>% addPolygons(data = region_shp, label = region_shp$density,
                  fillColor = myPal(region_shp$density), fillOpacity = 0.7,
                  color = "white", opacity = 1, weight = 2) %>% 
  addCircleMarkers(data = human_data, lat = ~y, lng = ~x, stroke = T,
                   fillColor = myPal2(human_data$date_decimal), fillOpacity = 0.7,
                   radius = 4, weight = 2, color = "black", opacity = 0.8)
```

<!----------------------------------------------------------------------------->
## Solution 2.2d {.smaller}

```{r, fig.height=3}
human_data <- map_data[map_data$species == "human",]
myPal <- colorQuantile(palette = "YlOrRd", domain = region_shp$density)
myPal2 <- colorNumeric(palette = "viridis", domain = human_data$date_decimal)
m %>% addPolygons(data = region_shp, label = region_shp$density,
                  fillColor = myPal(region_shp$density), fillOpacity = 0.7,
                  color = "white", opacity = 1, weight = 2) %>% 
  addCircleMarkers(data = human_data, lat = ~y, lng = ~x,
                   fillColor = myPal2(human_data$date_decimal), fillOpacity = 0.7, radius = 4,
                   stroke = T, weight = 2, color = "black", opacity = 0.8)
```

<!----------------------------------------------------------------------------->
## Add labels with some html formatting {.smaller}

```{r}
labels <- sprintf("<strong>%s</strong><br/>%g people / km<sup>2</sup>",
                   region_shp$Region_Nam, round(region_shp$density, 0)) %>%
  lapply(htmltools::HTML)

m %>% addPolygons(data = region_shp, fillColor = myPal(region_shp$density), label = labels,
                  weight = 2, opacity = 1, color = "white", fillOpacity = 0.7,
                  highlightOptions = highlightOptions(color = "black", weight = 3,
                                                      bringToFront = TRUE, opacity = 1))
```

<!----------------------------------------------------------------------------->
# Using leaflet and environmental data in raster format

<!----------------------------------------------------------------------------->
## Loading raster data

Load data using raster() from the **raster** package

```{r, eval = T}
density <- raster::raster("../../Day 1/data/HumanPopulation.grd")
density
```

<!----------------------------------------------------------------------------->
## Plotting raster data

We can use the base plotting function ``plot()`` on the object we've created

```{r, eval = T, fig.height = 4}
plot(density)
```

<!----------------------------------------------------------------------------->
## Plotting raster data

Logging density will provide more texture to map

```{r, eval = T, fig.height = 4}
plot(log(density))
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet

We use ``colorNumeric()`` again, as we did to colour by date earlier

```{r, eval = T}
colour_pal <- c("#FFFFCC", "#41B6C4", "#0C2C84")
raster::values(density) %>% range(na.rm = T)
myPal <- colorNumeric(palette = colour_pal, domain = c(0, 30000), na.color = "red")

m <- leaflet() %>%
  addRasterImage(x = density, colors = myPal, opacity = 0.8)
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet

```{r, eval = T}
m
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

Let's log density as we did before

```{r, eval = T}
raster::values(density) %>% range(na.rm = T) %>% log
myPal <- colorNumeric(palette = colour_pal, domain = c(-1.5, 11),
                      na.color = "transparent")
```

We have to exclude some data, as there are 0 density polygons and log(0) gives -infinity

```{r, eval = T}
sum(raster::values(density) > exp(-1.5), na.rm = T) /
  sum(raster::values(density) >= 0, na.rm = T)
```


<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}


Plot the logged raster data with some tiles and a legend for the colour scheme.

```{r, eval = T}
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.Positron) %>%
  addRasterImage(x = log(density), colors = myPal, opacity = 0.8) %>%
  addLegend(pal = myPal, values = c(-1.5, 11), opacity = 0.8,
            title = "Human<br>population<br>density")
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

```{r, eval = T, echo = F}
m
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}


Instead of using ``addProviderTiles(provider = providers$CartoDB.Positron)``

We order things here so that we plot

1) The background tiles only (``providers$CartoDB.PositronNoLabels``)
2) Then the raster data (``addRasterImage()``)
3) and finally the labels (``providers$CartoDB.PositronOnlyLabels``)

```{r, eval = T, warning=F}
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.PositronNoLabels) %>%
  addRasterImage(x = log(density), colors = myPal, opacity = 0.8) %>%
  addProviderTiles(provider = providers$CartoDB.PositronOnlyLabels) %>%
  addLegend(pal = myPal, values = c(-1.5, 11), opacity = 0.8,
            title = "Human<br>population<br>density")
```

<!----------------------------------------------------------------------------->
## Plotting raster data with leaflet {.smaller}

```{r, eval = T, echo = F}
m
```

<!----------------------------------------------------------------------------->
## Your turn 2.2e

Adapt the code below to add region outlines that highlight when the cursor hovers an area. Use ``addPolygon()`` with ``fillOpacity = 0``

```{r, eval = T, message = F}
colour_pal <- c("#FFFFCC", "#41B6C4", "#0C2C84")
myPal <- colorNumeric(palette = colour_pal, domain = c(-1.5, 11),
                      na.color = "transparent")

m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.PositronNoLabels) %>%
  addRasterImage(x = log(density), colors = myPal, opacity = 0.8) %>%
  addProviderTiles(provider = providers$CartoDB.PositronOnlyLabels) %>%
  addLegend(pal = myPal, values = c(-1.5, 11), opacity = 0.8,
            title = "Human<br>population<br>density")

```


<!----------------------------------------------------------------------------->
## Solution 2.2e {.smaller}

```{r, eval = T, message = F, warning=F}
m <- leaflet() %>%
  addProviderTiles(provider = providers$CartoDB.PositronNoLabels) %>%
  addRasterImage(x = log(density), colors = myPal, opacity = 0.8) %>%
  addProviderTiles(provider = providers$CartoDB.PositronOnlyLabels) %>%
  addLegend(pal = myPal, values = c(-1.5, 11), opacity = 0.8,
            title = "Human<br>population<br>density") %>% 
  addPolygons(data = region_shp, label = region_shp$Region_Nam, fillOpacity = 0,
              color = "black", weight = 1, opacity = 0.6,
              highlightOptions = highlightOptions(color = "white", weight = 3,
                                                  bringToFront = TRUE, opacity = 1))
  
```

<!----------------------------------------------------------------------------->
## Solution 2.2e

```{r, eval = T, echo = F, message = F, fig.height=4.5}
m
```

<!----------------------------------------------------------------------------->
# Exporting leaflet map

<!----------------------------------------------------------------------------->
## Exporting as a widget

``saveWidget()`` from the **htmlwidgets** package can save the map as a .html file.

You can email this to someone, add to a webpage etc.

There is no need for R to be running in the background.

```{r, eval=F}
saveWidget(widget = m, file = "filename.html")
```

<!----------------------------------------------------------------------------->
## Exporting as an image

``mapshot()`` from the **mapview** package can save the map as a static image in either a .png, .pdf, or .jpeg file.

```{r, eval=F}
m <- m %>% setView(m, lat = -6.4333, lng = 38.9, zoom = 7.5)
mapshot(m, file = "filename.png", width = 7, height = 6)

```

<!----------------------------------------------------------------------------->
## Exporting as an image

``mapshot()`` from the **mapview** package can save the map as a static image in either a .png, .pdf, or .jpeg file.

Combine with ``setView()`` from the **leaflet** package to control where map view is centered and level of zoom

```{r, eval=F}
m <- m %>% setView(lat = -6.4333, lng = 38.9, zoom = 7.5)
mapshot(m, file = "filename_zoom.png", width = 7, height = 6)
```
