---
title: "Data Visualisation using RShiny"
author: "Laurie Baker, Elaine Ferguson, Will Harvey and Rachel Steenson"
date: "**Tanzania**, August 2019"
output: 
  ioslides_presentation:
    widescreen: true
runtime: shiny
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(leaflet)
library(lubridate)
library(dplyr)
library(rgdal)
library(rgeos)
library(shinyWidgets)

# Load in the raw data
raw_data <- read.csv("data/raw_data.csv", stringsAsFactors=FALSE)

# Add year and decimal date to data
leaflet_data <- raw_data %>% 
  mutate(year = substr(date, 1,4), date=ymd(date), date_decimal = decimal_date(date)) 

## Create a colour palette for points
palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D")

## Load region shapefile
regions <- readOGR("data/TZ_Region_2012","TZ_Region_2012")

# Get the unique names of the species for the drop down menu
all_species <- unique(leaflet_data$species)


```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<style>
div.footnotes {
position: absolute;
bottom: 0;
margin-bottom: 10px;
width: 80%;
font-size: 0.6em;}
.blue {color: #4E23B8;}
.ui-background {background-color: #fcf0cd;}
.server-background {background-color: #dfecfc;}
</style>

<script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script>
<script>
$(document).ready(function() {
$('slide:not(.backdrop):not(.title-slide)').append('<div class=\"footnotes\">');

$('footnote').each(function(index) {
var text  = $(this).html();
var fnNum = (index+1).toString();
$(this).html(fnNum.sup());

var footnote   = fnNum + '. ' + text + '<br/>';
var oldContent = $(this).parents('slide').children('div.footnotes').html();
var newContent = oldContent + footnote;
$(this).parents('slide').children('div.footnotes').html(newContent);
});
});
</script>

<!----------------------------------------------------------------------------->
## Course Overview {.smaller}

<h3><b>Day 1:</b></h3>
<ul>
<li> 1.1 Getting to know your data </li>
<li> 1.2 Data subsetting and summarising </li>
<li> 1.3 Build exploratory plots </li>
<li> 1.4 Building an interactive plot in RShiny </li>
</ul>

<h3><b>Day 2:</b></h3>
<ul>
<li> 2.1 Introduction to leaflet </li>
<li> 2.2 Building a leaflet map in R </li>
<li> <b class="blue"> 2.3 Build an interactive map in RShiny </b></li>
</ul>

<h3><b>Day 3:</b></h3>
<ul>
<li> 3.1 Review </li>
<li> 3.2 Build your own apps! </li>
</ul>

<!----------------------------------------------------------------------------->

## Interacting with leaflet maps in Rshiny
  
Some interactive elements (e.g. pop-ups, zoom) are available in leaflet itself.

However, if we want to interactively change which data points or layers will be displayed on the map, we need to bring it into Shiny.

We can use many of the same widgets used to interact with regular plots to interact with leaflet maps!
  

<!----------------------------------------------------------------------------->

## A simple leaflet map to be made interactive

``` {r, echo=TRUE}

## Initialise map with tile. Set view and zoom.
m <- leaflet(width=800, height=300) %>% 
  addProviderTiles("Stamen.Terrain") %>%
  setView(c(gCentroid(regions)@coords)[1], c(gCentroid(regions)@coords)[2], 
          zoom = 5)
m
  


```


<!----------------------------------------------------------------------------->

## A simple leaflet map to be made interactive

``` {r, echo=TRUE, eval=TRUE}

## Add region shapefile
m <- m %>% 
  addPolygons(data=regions,color="black",fillColor = "white",
              label=regions$Region_Nam, weight=1, fillOpacity=0.7)
m


```

<!----------------------------------------------------------------------------->

## A simple leaflet map to be made interactive

``` {r, echo=FALSE, eval=TRUE}


## Add points coloured by species with legend
popupInfo <- 
  paste("Date: ", leaflet_data$date, "<br>",
        "Species: ", leaflet_data$species, "<br>",
        "Age: ", leaflet_data$age, "<br>",
        "Sex: ", leaflet_data$sex, "<br>",
        sep = " ")

pal <- colorFactor(palette, domain = sort(unique(leaflet_data[,"species"])))


```


``` {r, echo=TRUE, eval=TRUE}



m %>% 
  addCircles(data=leaflet_data,lng=~leaflet_data$x,lat=~leaflet_data$y,
             color = pal(leaflet_data[,"species"]), weight=2,
             opacity=1, fillOpacity=1, popup = popupInfo) %>%
  addLegend(position = "bottomright", title = "Species",
            pal = pal, values = leaflet_data[,"species"], opacity=1,
            labFormat = labelFormat(big.mark = "")) 

```

<!----------------------------------------------------------------------------->

## What widgets could we add to make the map more interactive?

- ``selectInput()`` to select alternative variables to colour points by 
- ``sliderInput()`` to choose date range of plotted points 
- ``checkboxGroupInput`` to select which background polygon layers to display 
- ``pickerInput()`` to choose a subset of the species to plot (from the <b>shinyWidgets</b> package) 
- ... 


<!----------------------------------------------------------------------------->

## Outputting a leaflet map in Shiny

- ``renderLeaflet()`` is used in the <b>server</b> to create the leaflet map 
- ``leafletOutput()`` is used in the <b>ui</b> to diplay the map


<!----------------------------------------------------------------------------->

## Create a map app with choice of variable to colour by

<b>shinyUI Side Panel</b>
```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
selectInput(inputId="colourby", label="Colour Cases By:",
            choices = c("species","date"), selected="species")

```

<b>shinyUI Main Panel</b>
```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
leafletOutput("mymap",width=1000,height=700)

```

<!----------------------------------------------------------------------------->
## Create a map app with choice of variable to colour by

<b>shinyServer ``reactive`` section</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}

pal <- reactive({
  colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
  if(input$colourby == "species"){ 
    colorFactor(palette, domain = sort(unique(leaflet_data[,colourby_col])))  
  }else if(input$colourby == "date"){
    colorNumeric(palette, range(leaflet_data[,colourby_col]))
  }
})
  
```


<!----------------------------------------------------------------------------->
## Create a map app with choice of variable to colour by {.smaller}

<b>shinyServer ``renderLeaflet()`` section</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}
output$mymap <- renderLeaflet({

  colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")

  leaflet() %>% 
    addProviderTiles("Stamen.Terrain") %>%
    setView(c(gCentroid(regions)@coords)[1], c(gCentroid(regions)@coords)[2], zoom = 6) %>% 
    addPolygons(data=regions,color="black",fillColor = "white",
                label=regions$Region_Nam, weight=1, fillOpacity=0.7) %>%
    addCircles(data=leaflet_data,lng=~leaflet_data$x,lat=~leaflet_data$y,
               color = pal()(leaflet_data[,colourby_col]),
               opacity=1, fillOpacity=1, popup = popupInfo) %>%
    addLegend(position = "bottomright", title = input$colourby,
              pal = pal(), values = leaflet_data[,colourby_col], opacity=1,
              labFormat = labelFormat(big.mark = "")) 
})


```

<!----------------------------------------------------------------------------->
## Create a map app with choice of variable to colour by 

```{r, echo=FALSE}

shinyApp(
  
  ui=shinyUI(fluidPage(
    
    sidebarLayout(
      
      # Sidebar containing the widgets
      sidebarPanel(
        
        # Drop down menu to choose variable by which points will be coloured
        selectInput(inputId="colourby", label="Colour Cases By:",
                    choices = c("species","date"), selected="species")
        
      ),
      
      
      # Show a plot of the map
      mainPanel(
        leafletOutput("mymap",width=580,height=450)
      )
      
    )
    
  
  )),
  
  server = shinyServer(function(input, output) {
  
  
    # Get point colours based on chosen variable
    pal <- reactive({
      colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
      if(input$colourby == "species"){ 
        colorFactor(palette, domain = sort(unique(leaflet_data[,colourby_col])))  
      }else if(input$colourby == "date"){
        colorNumeric(palette, range(leaflet_data[,colourby_col]))
      }
    })
    
    
    ## Render map
    output$mymap <- renderLeaflet({
      
      ## Initialise map with tile. Set central point of viewing window and initial amount of zoom.
      m <- leaflet() %>% 
        addProviderTiles("Stamen.Terrain") %>%
        setView(c(gCentroid(regions)@coords)[1], c(gCentroid(regions)@coords)[2], zoom = 5)
      
      
      ## Add region shapefile
      m <- m %>% 
        addPolygons(data=regions,color="black",fillColor = "white",
                    label=regions$Region_Nam, weight=1, fillOpacity=0.7)
      
      ## Add coloured points and legend
      colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
      m %>% 
        addCircles(data=leaflet_data,lng=~leaflet_data$x,lat=~leaflet_data$y,
                   color = pal()(leaflet_data[,colourby_col]),
                   opacity=1, fillOpacity=1, popup = popupInfo) %>%
        addLegend(position = "bottomright", title = input$colourby,
                  pal = pal(), values = leaflet_data[,colourby_col], opacity=1,
                  labFormat = labelFormat(big.mark = "")) 
      
    })
    
  })

  
)


```

<!----------------------------------------------------------------------------->
## Your turn  

Complete section 2.3a of the handout.<br><br>



<!----------------------------------------------------------------------------->
## Summary - section 2.3a  

<b>shinyUI Side Panel</b>
```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
selectInput(inputId="colourby", label="Colour Cases By:",
            choices = c("species","date","sex","age"),
            selected="species")

```


<b>shinyServer ``reactive`` section</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}

pal <- reactive({
  colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
  if(input$colourby %in% c("species","sex")){ 
    colorFactor(palette, domain = sort(unique(leaflet_data[,colourby_col])))  
  }else if(input$colourby %in% c("date","age")){
    colorNumeric(palette, range(leaflet_data[,colourby_col]))
  }
})
  
```


<!----------------------------------------------------------------------------->
## The shinyWidgets package {.smaller}

- provides an extension of the widgets available in base R shiny
- many provide similar basic functions to those we've used already, but are more customisable or have added functionality
- these widgets can be explored in the shinyWidgets online gallery
- e.g. ``pickerInput``

```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
pickerInput(inputId = "species", label = "Species:", 
            choices = sort(all_species), selected= all_species, 
            options = list(`actions-box` = TRUE,`live-search` = TRUE), multiple = T),

```


```{r, echo=FALSE, eval=TRUE}



shinyApp(
  ui=shinyUI(fluidPage(
      pickerInput(inputId = "species", label = "Pick species:",
                  sort(all_species), selected= all_species, 
                  options = list(`actions-box` = TRUE,`live-search` = TRUE), multiple = T)
    )),
  server=shinyServer(function(input, output){})
  
)

```

  
<!----------------------------------------------------------------------------->
## Subsetting the data to be displayed

<b>shinyServer new ``reactive`` sections</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}

## Subset data based on inputs
leaflet_data_sub<- reactive({
  leaflet_data %>% 
    filter(species %in% input$species)
})

## Create text pop-up information for each point in subsetted data
popupInfo <- reactive({
  paste("Date: ", leaflet_data_sub()$date, "<br>",
        "Species: ", leaflet_data_sub()$species, "<br>",
        "Age: ", leaflet_data_sub()$age, "<br>",
        "Sex: ", leaflet_data_sub()$sex, "<br>",
        sep = " ")
})

```
  

<!----------------------------------------------------------------------------->
## Subsetting the data to be displayed

<b>shinyServer ``renderLeaflet()`` changes</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}
addCircles(data=leaflet_data_sub(),lng=~leaflet_data_sub()$x,lat=~leaflet_data_sub()$y,
           color = pal()(leaflet_data_sub()[,colourby_col]),
           opacity=1, fillOpacity=1, popup = popupInfo()) %>%
addLegend(position = "bottomright", title = input$colourby,
          pal = pal(), values = leaflet_data_sub()[,colourby_col], opacity=1,
          labFormat = labelFormat(big.mark = "")) 
})


```

<!----------------------------------------------------------------------------->
## Subsetting the data to be displayed

```{r, echo=FALSE,eval=TRUE, class.source="server-background"}

shinyApp(
  ui = shinyUI(fluidPage(
  
  sidebarLayout(
    
    # Sidebar containing the widgets
    sidebarPanel(
      
      # Drop down menu to choose variable by which points will be coloured
      selectInput(inputId="colourby", label="Colour Cases By:",
                  choices = c("species","date","sex","age"),
                  selected="species"),
      
      br(),
      
      # Menu for selecting which species to display
      pickerInput(inputId = "species", label = "Species:",
                  sort(all_species), selected= all_species, # Use sort to get names in alphabetical order
                  options = list(`actions-box` = TRUE,`live-search` = TRUE), multiple = T)
      
      
    ),
    
    
    # Show a plot of the map
    mainPanel(
      leafletOutput("mymap",width=620,height=490)
    )
    
  )
  

))
,
  server=shinyServer(function(input, output) {
  
  ## Subset data based on inputs
  leaflet_data_sub<- reactive({
    leaflet_data %>% 
      filter(species %in% input$species)
    
  })
  
  ## Create text pop-up information for each point in subsetted data
  popupInfo <- reactive({
    paste("Date: ", leaflet_data_sub()$date, "<br>",
          "Species: ", leaflet_data_sub()$species, "<br>",
          "Age: ", leaflet_data_sub()$age, "<br>",
          "Sex: ", leaflet_data_sub()$sex, "<br>",
          sep = " ")
  })
  
  # Get point colours based on chosen variable
  pal <- reactive({
    colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
    if(input$colourby %in% c("species","sex")){ 
      colorFactor(palette, domain = sort(unique(leaflet_data[,colourby_col])))  
    }else if(input$colourby %in% c("date","age")){
      colorNumeric(palette, range(leaflet_data[,colourby_col]))
    }
  })
  
  
  ## Render map
  output$mymap <- renderLeaflet({
    
    ## Initialise map with tile. Set central point of viewing window and initial amount of zoom.
    m <- leaflet() %>% 
      addProviderTiles("Stamen.Terrain") %>%
      setView(c(gCentroid(regions)@coords)[1], c(gCentroid(regions)@coords)[2], zoom = 6)
    
    
    ## Add region shapefile
    m <- m %>% 
      addPolygons(data=regions,color="black",fillColor = "white",
                  label=regions$Region_Nam, weight=1, fillOpacity=0.7)
    
    ## Add coloured points and legend
    colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
    m %>% 
      addCircles(data=leaflet_data_sub(),lng=~leaflet_data_sub()$x,lat=~leaflet_data_sub()$y,
                 color = pal()(leaflet_data_sub()[,colourby_col]),
                 opacity=1, fillOpacity=1, popup = popupInfo()) %>%
      addLegend(position = "bottomright", title = input$colourby,
                pal = pal(), values = leaflet_data_sub()[,colourby_col], opacity=1,
                labFormat = labelFormat(big.mark = "")) 
    
  })
  
})
)

```


<!----------------------------------------------------------------------------->
## Your turn  

Complete section 2.3b of the handout.<br><br>


<!----------------------------------------------------------------------------->

## Summary - section 2.3b  

<b>Above shinyUI</b>
```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
all_regions <- unique(leaflet_data$region)
```

<b>shinyUI Side Panel</b>
```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
pickerInput(inputId = "region", label = "Region:",
            sort(all_regions), selected= all_regions, 
            options = list(`actions-box` = TRUE,`live-search` = TRUE), multiple = T)
```

<b>shinyServer ``reactive`` section</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}
leaflet_data_sub<- reactive({
  leaflet_data %>% 
    filter(region %in% input$region)
})
```

<!----------------------------------------------------------------------------->

## Changing the displayed map layers

- As you add more map layers (shapefiles, rasters and points), your map can become overcrowded and take a long time to load
- Allowing the user to select the layers they're interested in can make your app faster and visually clearer

<b>shinyUI Side Panel</b>
```{r, echo=TRUE, eval=FALSE, class.source="ui-background"}
# Checkbox for choosing shapefiles to be displayed 
checkboxInput("shapefile", label = "Region shapefile:", value = TRUE)
```

<!----------------------------------------------------------------------------->

## Changing the displayed map layers {.smaller}
<b>shinyServer ``leaflet`` code</b>
```{r, echo=TRUE, eval=FALSE, class.source="server-background"}

## Initialise map with tile. Set viewing window and initial zoom.
m <- leaflet() %>% addProviderTiles("Stamen.Terrain") %>%
  setView(c(gCentroid(regions)@coords)[1], c(gCentroid(regions)@coords)[2], zoom = 6)

## Add region shapefile if box is checked
if(input$shapefile == TRUE){
  m <- m %>% addPolygons(data=regions,color="black",fillColor = "white",
                         label=regions$Region_Nam, weight=1, fillOpacity=0.7)}

## Add coloured points and legend
colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
m %>% addCircles(data=leaflet_data_sub(),lng=~leaflet_data_sub()$x,lat=~leaflet_data_sub()$y,
                 color = pal()(leaflet_data_sub()[,colourby_col]),
                 opacity=1, fillOpacity=1, popup = popupInfo()) %>%
      addLegend(position = "bottomright", title = input$colourby,
                pal = pal(), values = leaflet_data_sub()[,colourby_col], opacity=1,
                labFormat = labelFormat(big.mark = "")) 
```

- in an ``if(){}`` statement, the code inside the curly brackets is only evaluated if the conditions in the round bracket are met


<!----------------------------------------------------------------------------->

## Changing the displayed map layers {.smaller}

```{r, echo=FALSE, eval=TRUE, class.source="server-background"}
shinyApp(
  ui=shinyUI(fluidPage(
  
  sidebarLayout(
    
    # Sidebar containing the widgets
    sidebarPanel(
      
      # Drop down menu to choose variable by which points will be coloured
      selectInput(inputId="colourby", label="Colour Cases By:",
                  choices = c("species","date","sex","age"),
                  selected="species"),
      
      br(),
      
      # Menu for selecting which species to display
      pickerInput(inputId = "species", label = "Species:",
                  sort(all_species), selected= all_species, # Use sort to get names in alphabetical order
                  options = list(`actions-box` = TRUE,`live-search` = TRUE), multiple = T),
      
      br(),
      
      # Checkbox for selecting whether region shapefile should be displayed 
      checkboxInput("shapefile", label = "Region shapefile:", value = TRUE)
      
      
    ),
    
    
    # Show a plot of the map
    mainPanel(
      leafletOutput("mymap",width=630,height=490)
    )
    
  )
  

)),

server = shinyServer(function(input, output) {
  
  ## Subset data based on inputs
  leaflet_data_sub<- reactive({
    leaflet_data %>% 
      filter(species %in% input$species)
    
  })
  
  ## Create text pop-up information for each point in subsetted data
  popupInfo <- reactive({
    paste("Date: ", leaflet_data_sub()$date, "<br>",
          "Species: ", leaflet_data_sub()$species, "<br>",
          "Age: ", leaflet_data_sub()$age, "<br>",
          "Sex: ", leaflet_data_sub()$sex, "<br>",
          sep = " ")
  })
  
  # Get point colours based on chosen variable
  pal <- reactive({
    colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
    if(input$colourby %in% c("species","sex")){ 
      colorFactor(palette, domain = sort(unique(leaflet_data[,colourby_col])))  
    }else if(input$colourby %in% c("date","age")){
      colorNumeric(palette, range(leaflet_data[,colourby_col]))
    }
  })
  
  
  ## Render map
  output$mymap <- renderLeaflet({
    
    ## Initialise map with tile. Set central point of viewing window and initial amount of zoom.
    m <- leaflet() %>% 
      addProviderTiles("Stamen.Terrain") %>%
      setView(c(gCentroid(regions)@coords)[1], c(gCentroid(regions)@coords)[2], zoom = 6)
    
    
    ## Add region shapefile if box is checked
    if(input$shapefile == TRUE){
      m <- m %>% 
        addPolygons(data=regions,color="black",fillColor = "white",
                    label=regions$Region_Nam, weight=1, fillOpacity=0.7)}
    
    
    ## Add coloured points and legend
    colourby_col <- ifelse(input$colourby!="date",input$colourby,"date_decimal")
    m %>% 
      addCircles(data=leaflet_data_sub(),lng=~leaflet_data_sub()$x,lat=~leaflet_data_sub()$y,
                 color = pal()(leaflet_data_sub()[,colourby_col]),
                 opacity=1, fillOpacity=1, popup = popupInfo()) %>%
      addLegend(position = "bottomright", title = input$colourby,
                pal = pal(), values = leaflet_data_sub()[,colourby_col], opacity=1,
                labFormat = labelFormat(big.mark = "")) 
    
  })
  
})

)

```

